CXXC-finger protein 1 associates with FOXP3 to stabilize homeostasis and suppressive functions of regulatory T cells

Version 1

Abstract

Abstract

FOXP3-expressing regulatory T (Treg) cells play a pivotal role in maintaining immune homeostasis and tolerance, with their activation being crucial for preventing various inflammatory responses. However, the mechanisms governing the epigenetic program in Treg cells during their dynamic activation remain unclear. In this study, we demonstrate that CXXC finger protein 1 (CXXC1) interacts with the transcription factor FOXP3 and facilitates the regulation of target genes by modulating H3K4me3 deposition. Cxxc1 deletion in Treg cells leads to severe inflammatory disease and spontaneous T-cell activation, with impaired immunosuppressive function. As a transcriptional regulator, CXXC1 promotes the expression of key Treg functional markers under steady-state conditions, which are essential for the maintenance of Treg cell homeostasis and their suppressive functions. Epigenetically, CXXC1 binds to the genomic regulatory regions of Treg program genes in mouse Treg cells, overlapping with FOXP3 binding sites. Given its critical role in Treg cell homeostasis, CXXC1 presents itself as a promising therapeutic target for autoimmune diseases.

Introduction

Regulatory T (Treg) cells are a distinct subset of CD4+ T cells that play a critical role in maintaining immune homeostasis and self-tolerance by suppressing excessive or aberrant immune responses to foreign or self-antigens 1–3. These cells can be further categorized into thymus-derived regulatory T cells (tTreg cells), periphery-derived Treg cells (pTreg cells), and induced Treg cells (iTreg cells) 4. They uniquely express the transcription factor FOXP3, a member of the forkhead winged-helix family, which is essential for Treg cell lineage commitment and suppressive function 5,6. Deletion or mutation of the Foxp3 gene leads to a range of immunological disorders, including allergies, immunopathology, and autoimmune diseases in both mice and humans 7,8.

It is well established that FOXP3 recruits various cofactors to form complexes that either promote or repress the expression of downstream genes, with histone and DNA modifications playing pivotal roles in this process. FOXP3 can activate or repress the transcription of key regulators of Treg cell activation and function by recruiting the histone acetyltransferases (HATs) or histone deacetylases (HDACs)9,10. Notably, FOXP3-bound sites exhibit significant enrichment of H3K27me3, a modification essential for FOXP3-mediated repressive chromatin remodeling under inflammatory conditions11. However, the direct role of FOXP3 as a transcriptional activator through interactions with epigenetic regulators, particularly via modulation of H3K4 trimethylation, remains poorly documented.

In mammals, six proteins have been identified that catalyze H3K4 methylation. These proteins contain the SET domain and include MLL1 (KMT2A), MLL2 (KMT2B), MLL3 (KMT2C), MLL4 (KMT2D), SETD1A, and SETD1B 12,13. For example, MLL1 plays a critical role as an epigenetic regulator in Treg cell activation and functional specialization14. Additionally, Placek et al. demonstrated that MLL4 is essential for Treg cell development by catalyzing H3K4me1 at distant unbound enhancers through chromatin looping 15. H3K4me3, which is enriched at the transcription start site (TSS) and the CpG island (CGI), converts chromatin into active euchromatin by recruiting activating factors 16. CXXC finger protein 1(CXXC1, also known as CFP1), which contains a SET1 interaction domain (SID), is required for binding to the histone H3K4 methyltransferases SETD1A and SETD1B 17. Previous studies have shown that CXXC1 regulates promoter patterns during T cell maturation, GM-CSF-derived macrophage phagocytosis, TH17 cell differentiation, and the function of ILC3 cells with aging by modulating H3K4me3 modification18–21. Despite the well-documented role of CXXC1 in various immune effector cells, its role in Treg cells remains unclear.

Here, we demonstrate that CXXC1 interacts with FOXP3 and enhances the expression of FOXP3 target genes. Treg cell-specific deletion of Cxxc1 triggers systemic autoimmunity, accompanied by multiorgan inflammation, ultimately resulting in early-onset fatal inflammatory disease in mice. Cxxc1-deficient Treg cells exhibit a significant disadvantage in proliferation and homeostasis, even in non-inflammatory mice where coexisting wild-type (WT) Treg cells were present. Moreover, Cxxc1-deficient Treg cells display intrinsic defects in the expression of key suppression molecules, including CTLA-4, CD25, ICOS, and GITR. Consistent with these findings, mechanistic studies strongly suggest that CXXC1 functions as an essential cofactor of FOXP3, playing a crucial role in sustaining H3K4me3 modifications at key Treg cell genes.

Materials and methods

Mice

All mice used in this study were bred for a minimum of seven generations on a C57BL/6 background. Mouse experiments, or cells from mice of the same genotype, compared littermates or age-matched control animals. The Cxxc1fl/fl mouse strain has been previously described 22. The Foxp3cre mice (JAX,016959) were generously provided by Bin Li (Shanghai Jiao Tong University School of Medicine, Shanghai, China). CD45.1 (NM-KI-210226) mice were purchased from the Nanjing Biomedical Research Institute of Nanjing University. Rag1-/- mice (stock#T004753) were purchased from GemPharmatech. 2D2 (MOG35-55-specific TCR transgenic) mice were graciously supplied by Prof. Linrong Lu (Zhejiang University School of Medicine, Hangzhou, Zhejiang, China). Foxp3cre (WT) and Foxp3cre Cxxc1fl/fl (cKO) were utilized at 3 weeks old unless otherwise stated. All mice were housed in the Zhejiang University Laboratory Animal Center under specific pathogen-free conditions, and all animal experimental procedures were approved by the Zhejiang University Animal Care and Use Committee (approval no.ZJU20230246).

Cell culture

HEK 293T cells were obtained from ATCC, and Plat E cells were kindly provided by Prof. Xiaolong Liu (Shanghai Institutes for Biological Sciences). Both cell lines were cultured in Dulbecco-modified Eagle medium (DMEM) containing 10% (v/v) fetal bovine serum (FBS), supplemented with 1% penicillin/streptomycin.

Immunofluorescence (IF) Staining

As previously described 23, coverslips were treated with a 0.01% poly-L-lysine solution (P4707; Sigma) for 10 minutes, air-dried, and then coated with CD4+ YFP+ Treg cells. The cells were then fixed in 4% formaldehyde for 15 minutes at room temperature, permeabilized with 0.2% Triton X-100, and blocked with 1% BSA. Antibodies against CXXC1 (ab198977; Abcam) and FOXP3 (17-5773-82; Invitrogen) were diluted in Image iT FX signal enhancer (I3693; Invitrogen) and incubated with cells overnight at 4 °C. After washing with phosphate-buffered saline (PBS), the cells were incubated with a goat anti-rabbit antibody Alexa Fluor 594 (1:250;10015289; Invitrogen) secondary antibody and stained with DAPI (200 ng/mL; D523; Dojindo). Slides were washed with PBS and sealed with an antifade solution(P36934; Invitrogen) before imaging with an Olympus FV3000 fluorescence microscope. The images were visualized using the FV31-SW software.

Co-immunoprecipitation and western blot

Harvest the appropriate transfected cell lines and primary cells from culture and wash them with ice-cold PBS. Lyse the cells in NETN300 buffer (300 mM NaCl, 0.5 mM EDTA, 0.5% (v/v) NP-40, 20 mM Tris-HCl pH 8.0) supplemented with a protease inhibitor (1:100, P8340, Sigma-Aldrich) and PMSF (1 mM) on ice for 10 minutes. Take a small portion of the whole-cell lysate as input, and incubate the remaining lysate with either Anti-FLAG M2 Beads (M8823; Sigma) or Anti-HA Beads (HY-K0201; MCE) on a rotator at 4°C overnight. For the endogenous Co-IP assay targeting CXXC1 and FOXP3, incubate the cell lysate with protein G magnetic beads along with anti-CXXC1 (ab198977; Abcam) or anti-FOXP3 (14-4774-82; Invitrogen) antibodies on a rotator at 4°C overnight. Wash the beads three times with IP buffer (100 mM NaCl, 0.5 mM EDTA, 0.5% (v/v) NP-40, 20 mM Tris-HCl pH 8.0) to remove non-specific binding. Boil the washed beads with 1× Laemmli sample buffer (1610747; Bio-Rad) to elute the bound proteins. Separate the denatured proteins by SDS-PAGE. Transfer the separated proteins onto PVDF membranes (Millipore) for immunoblotting.Immunoblot the PVDF membranes (IPVH00010) with the following antibodies: anti-CXXC1 (1:1000; ab198977; Abcam), anti-FOXP3 (1:500; 14-7979-80; Invitrogen), anti-FlAG (1:1000; 14793; Cell Signaling Technology), anti-HA (1:1000; 3724S; Cell Signaling Technology).Detect the immunoblotted proteins using a secondary HRP-conjugated goat anti-rabbit antibody (1:1000; HA1001-100; Huabio) and visualize the bands using an appropriate detection method.

Flow cytometry

Spleen and lymph nodes were harvested from 3-week-old mice. Tissues were placed in a dish containing PBS with 2% fetal bovine serum (FBS). The tissues were minced into small pieces and then filtered through a 70 µM nylon filter to obtain a single-cell suspension. For surface marker analysis, cells were incubated for 15 minutes with purified anti-mouse CD16/32 antibody (101320; BioLegend) to block Fc receptors. After blocking, cells were stained with the indicated antibodies for surface markers. To determine cytokine expression, cells were stimulated for 4 hours at 37℃ with phorbol12-myristate13-acetate (50ng/mL; S1819; Beyotime), ionomycin (1mg/ml; S1672; Beyotime), and brefeldin A (BFA;00-4506-51; Invitrogen). After stimulation, cells were labeled with a fixable viability dye and surface markers. Cells were then fixed and permeabilized according to the manufacturer’s instructions (00-8222-49; Invitrogen). For transcription factor staining, samples were fixed using the FOXP3/Transcription Factor Staining Buffer Set (00-5523-00; Invitrogen). Flow cytometry was performed using a BD Fortessa (BD Biosciences). Flow cytometry data were acquired and analyzed using FlowJo software.

The following antibodies were purchased from Invitrogen or BioLegend: Zombie Violet f-ixable viability(423113), Zombie NIR fixable(423105), CD25(PC61), CD8α(53-6.7), CD62L(MEL-14), PD-1(J43), CD44(IM7), IL-17A(TC1118H10), KL-RG1(2F1), CD4 (GK1.5), TCRβ(H57-597), IFN-γ(XMG1.2), FOXP3(FJK-16s), CTLA-4(UC10-4B9), ICOS(15F9), GITR(DTA-1), LAG3(C9B7W), TCRVβ11(RR3-15), CD45RB(C363-16A) CD69 (H1.2F3), MHCII(M5/114.15.2), CD80(16-10A1), T-bet(eBio4B10), IKZF4(ESB7C2), IL-4(11B11), CCR7(4B12), CD73(TY/11.8).

Real-time PCR

Total RNA was extracted from Treg cells using the RNAAiso Plus(9109; Takara) reagent according to the manufacturer’s instructions, and cDNA synthesis was performed using the Prime Script RT Reagent Kit (Takara). TB Green Premix Ex Taq (RR420A; Takara) was used for quantitative real-time PCR (qPCR). The expression levels of target mRNA were normalized to the level of β-actin expression. The primers for qPCR are as follows:

Cxxc1 qPCR Forward: ATCCGGGAATGGTACTGTCG

Cxxc1 qPCR Reverse: CTGTGGAGAAGATTTGTGGG

β-actin qPCR Forward: CTGTCCCTGTATGCCTCTG

β-actin qPCR Reverse: ATGTCACGCACGATTTCC

CD4+T and YFP+ Treg cells adoptive transfer in Experimental autoimmune encephalomyelitis (EAE)

CD4+YFP+ Treg cells from Foxp3cre and Foxp3cre Cxxc1fl/fl mice were enriched using the Mouse CD4 T Cell Isolation Kit (480005; Biolegend) and then sorted using the BD Aria II flow cytometer. Naïve CD4+T cells from 2D2 (MOG35-55-specificTCR transgenic) mice were isolated using the Mouse CD4 Naïve T cell Isolation Kit480039 (480039; Biolegend). As previously described24, 2D2 naïve CD4+ T cells alone (5 x 105 per mouse), or 2D2 naïve CD4+ T cells (5x 105 per mouse) together with WT or Foxp3creCxxc1fl/fl Treg cells (2x105 per mouse), were transferred into Rag1−/− mice via the tail vein. One day after cell transfer, the recipient mice were inoculated subcutaneously (s.c.) with 200μg MOG35-55 peptide (MEVGWYRSPFSRVVHLYRNGK; GenemeSynthesis) emulsified in complete Freund’s adjuvant (CFA) (F5506; Sigma). Intravenous administration of 200 ng of Pertussis toxin (181; List Biological Laboratories) was performed on days 0 and 2 after peptide inoculation. The severity of EAE was monitored and blindly graded using a clinical score from 0 to 5: 0, no clinical signs; 1, limp tail; 2, paraparesis (weakness, incomplete paralysis of one or two hind limbs); 3, paraplegia (complete paralysis of two hind limbs); 4, paraplegia with forelimb weakness or paralysis; 5, dying or death.

Isolation lymphocytes from the central nervous system (CNS)

On day 14 after EAE induction, mice were perfused with transcardially administered PBS to eliminate contaminating blood cells in the central nervous system (CNS). The forebrain and cerebellum were dissected to expose the spinal cord, which was then carefully removed from the spinal canal. The fresh spinal cord was harvested and cut into 2 mm pieces. The CNS tissue pieces were homogenized using a syringe and passed through a 70 µM cell strainer to obtain a single-cell suspension. The single-cell suspension was digested with collagenase D (2μg/ml;11088858001; Roche) and deoxyribonuclease I (DNase I; 1 μg/ml; DN25; Sigma-Aldrich) at 37°C for 20 minutes under rotation. After digestion, the cell suspension was centrifuged to pellet the cells. The cell pellets were resuspended in 40% Percoll and layered onto a discontinuous Percoll gradient. Centrifugation at 80% Percoll allowed for the separation of cells at the 40–80% Percoll interface, which were collected as CNS mononuclear cells. The collected CNS mononuclear cells were washed with PBS to remove any remaining Percoll. CNS mononuclear cells were stimulated for 4 hours with PMA and ionomycin in the presence of Brefeldin A to induce cytokine production. After stimulation, cells were fixed, rendered permeable, and stained with appropriate antibodies for intracellular cytokine detection.

Histological analyses

The lungs, skin, liver, and colon were excised from three-week-old mice. Prior to histological analysis, the samples were fixed in formalin, embedded in paraffin, and stained with hematoxylin and eosin. For CNS histology, spinal cords were fixed in 4% paraformaldehyde, paraffin-embedded, sectioned, and stained with Luxol Fast Blue and hematoxylin and eosin (H&E). To examine colon histology, colons from Rag1−/− hosts were similarly processed and stained with H&E.

Adoptive transfer colitis model

Colitis was induced following the protocol described25. In brief, CD4+ YFP+ Treg cells were isolated from 3-week-old CD45.2+Foxp3creCxxc1fl/fl and CD45.2+ Foxp3cre mice. A total of 2x105 Treg cells from each group were mixed with 4x105 Teff cells (CD45.1+CD4+CD45RBhi) sorted from CD45.1+ mice and transferred into the Rag1-/- mice via intraperitoneal injection. Teff cells alone were transferred as a control group. Mouse body weight was measured weekly post-adoptive transfer. The percentage change in body weight was calculated by comparing the current weight with the initial weight on day 0. Mice were euthanized when any had reached 80% of their initial body weight. The large intestines were sectioned into 4μm thick slices and stained with hematoxylin.

In vitro Treg suppression assay

Naïve CD4+ T cells isolated from WT mice were labeled with CFSE (C34554; Invitrogen). CD4+YFP+ Treg cells from Foxp3cre and Foxp3creCxxc1fl/fl mice were cultured with naïve CD4+ T cells (1ⅹ105 cells) at various ratios in the presence of 2μg/mL anti-CD3(16-0031-85; Invitrogen) and 3μg/mL anti-CD28(16-0281-85; Invitrogen). On day 3, cells were analyzed by flow cytometry.

CUT&Tag

CUT&Tag assays of CD4+YFP+ Treg cells were conducted as previously described26. Briefly, approximately 1x105 single cells were carefully pipetted into wash buffer twice. The pelleted cells were resuspended in wash buffer, activated concanavalin (BP531; Bangs Laboratories), and incubated for 15 minutes at room temperature. Cells bound to the beads were resuspended in Dig-Wash Buffer and incubated with a 1:50 dilution of primary antibodies (rabbit anti-H3K4me3, Active Motif,39016; rabbit anti-CXXC1, abcam, ab198977; normal IgG, Cell Signaling, 2729) at 4 ℃ overnight. The beads were incubated with a secondary antibody (goat anti-rabbit IgG; SAB3700883; Sigma-Aldrish) diluted 1:100 in Dig-Wash buffer for 60 minutes at room temperature. Cells were treated with Hyperactive pG-Tn5 Transposase (S602; Vazyme) diluted in Dig-300 Buffer for 1 hour at room temperature. The cells were subsequently resuspended in Tagmentation buffer (10mM MgCl2 in Dig-300 Buffer) and incubated at 37℃ for 1 hour. To halt tagmentation, 10μl was spiked with 0.5M EDTA, 3μl with 10%SDS, and 3μl with 20mg/ml Proteinase K and incubated at 55℃ for 1h. DNA library amplification was performed according to the manufacturer’s instructions and purified using VAHTS DNA Clean Beads (N411; Vazyme). Libraries were sequenced on the Illumina NovaSeq platform (Annoroad Gene Technology).

CUT&Tag and ChIP-seq data analysis

FOXP3 ChIP-seq data was obtained from GSE121279. H3K27me3 ChIP-seq data was obtained from GSE14254. CUT&Tag and ChIP-seq reads were trimmed to 50 bp and aligned against the mouse genome build mm9 using Bowtie2(v2.3.4.1) with default parameters. All PCR duplicates and unmapped reads were removed. Peak calling was performed using MACS2(v2.1.1.20160309) and signal tracks for each sample were generated using the ‘wigToBigWig’ utility of UCSC. We classified the H3k4me3 peaks around TSSs into three groups: broad (>5 kb), medium (1-5kb), and narrow (<1 kb). The top 5% of the widest peaks were considered as broad peaks. The average intensity profiles were generated using deepTools (v2.5.4). Motif analysis was performed using the ‘findmotifsGenome.pl’ command inHomer2 package. Epigenetic factors were identified using the Epigenetic Factor Database(https://epifactors.autosome.org/) and then screened for those that exclusively regulate the expression of their target genes by modulating the deposition of H3K4me3.

Genomic distribution was analyzed using the “genomation” R package. GO pathway analysis was performed using the “clusterProfiler” R package. The sequencing information of CUT&Tag data used in this study is summarized in Supplementary Table S1.

Clustering analysis

Promoters were defined as ±2 kb regions flanking the annotated TSS. Reads in promoters were counted using the “coverage” command in bedtools (v2.26.0) and further normalized to RPKM. The k-means clustering of H3K4me3 and H3K27me3 enrichment at promoters was conducted using the “kmeans” function in R.

RNA-seq and data analysis

Total RNA was extracted from sorted CD4⁺YFP⁺ Treg cells using the RNeasy Plus Mini Kit (Qiagen, #74134), following the manufacturer’s protocol. RNA-Seq libraries were constructed and sequenced by Haplox (Nanchang, China), using an Illumina platform with paired-end reads of 150 bp. RNA-seq data was obtained from GSE82076. Raw reads were trimmed to 50 bp and mapped to the mouse genome (mm9) using TopHat (v2.1.1) with default parameters. Only uniquely mapped reads were kept for downstream analysis. The RNA abundance of each gene was quantified using Cufflinks (v2.2.1).

Single-Cell RNA-Sequencing

A total of 300,000 sort-purified CD4+YFP+ Treg cells from Foxp3cre and Foxp3cre Cxxc1fl/fl were resuspended in BD Pharmingen Stain Buffer (FBS) (554656; BD). Single cells were isolated using a chromium controller (BD platform, BD Bioscience) according to the manufacturer’s instructions, as previously described27. The single cells were labeled with sample tags using the BD Mouse Immune Single-Cell Multiplexing Kit (633793; BD). Following standard protocols, cDNA amplification and library construction were performed to generate scRNA-seq libraries.

Targeted scRNA-seq data processing

The raw FASTQ files were processed by BD Rhapsody using the Targeted analysis pipeline. After alignment and filtering, the distribution-based error correction(DBEC)-adjusted molecules were loaded into R studio(version 4.3.2). All subsequent analyses were performed using the package Seurat (version 4.4.0) with default parameters. Specifically, the scRNA-seq data counts were log-normalized. All targeted genes were scaled and then were used for Principal components analysis (PCA). The batch effects were removed by the HarmonyMatrix function in the Harmony package (version 1.1.0). The first 20 principal components were used to calculate nonlinear dimensionality reduction using RunUMAP. Differential gene expression (DEGs) between clusters was assessed using the FindAllMarkers function. The clusters were then annotated based on DEGs. Barplots were generated using ggplot2 (version 3.4.4). Heatmaps were generated using pheatmap (version 1.0.12).

Analysis of the single-cell TCR-seq repertoire

Raw V(D)J fastq reads were processed using BD Rhapsody Pipeline and then were analyzed using scRepertoire (version 1.12.0). The TCR clonotype was called using the nucleotide sequence of the CDR3 region for both TCR alpha and beta chains. For cells with multiple chains, the top two clonotypes with the highest expression were selected for downstream analysis. Clonal overlap between different cell types was calculated using the clonalOverlap function of scRepertoire. A clonotype was defined as expansion if it could be detected in at least two cells.

scRNA-seq trajectory analysis

UMAP embeddings obtained from the Seurat package were projected into the Slingshot(version 2.10.0) package to construct pseudotime Trajectories for Treg cells. Naïve subsets were set as the root state.

Whole genome bisulfite sequencing (WGBS) and data analysis

Sorted CD4+YFP+ Treg cells(3ⅹ106) were lysed in cell lysis buffer to release DNA. The bisulfite-treated DNA was used to prepare the sequencing library. DNA libraries were transferred to the Illumina Platform for sequencing using 150 bp paired-end reads. Raw reads were trimmed using TrimGalore (v0.4.4) with default parameters. Subsequently, the reads were mapped against the mm9 reference genome using Bismark v0.19.0 with parameters ‘--bowtie2’. PCR duplicates were removed and the methylation levels were calculated using ‘bismark_methylation_extractor’. We calculated the mean CpG methylation levels of various genome elements: promoter, 5′-UTR, exon, intron, 3′-UTR, genebody, intergenic, CGIs, and repeats using in-house scripts. The sequencing information of WGBS data used in this study is summarized in Supplementary Table S2.

Statistical analysis

The statistical significance analysis was performed using Prism 8.0 (GraphPad). Error bars are presented as mean ± SD. P values of < 0.05 were deemed statistically significant (*P < 0.05, **P < 0.01, ***P < 0.001 and ****P < 0.000). Statistical analyses were performed with unpaired Student’s t-test or two-way ANOVA and Holm–Sidak post hoc test.

Results

FOXP3 binds regulatory loci primed for activation and repression in Treg cells

FOXP3-mediated gene expression is well-recognized, with several studies highlighting its dual role as both a transcriptional activator and repressor 28–32. However, the regulation of Treg-specific gene transcription by FOXP3 through intermediary mechanisms, such as epigenetic factors, remains incompletely understood. To address this, we employed CUT&Tag to generate global H3K4me3 maps in Treg cells and compared these with an H3K27me3 ChIP-Seq dataset from Wei et al.33 at previously identified FOXP3-bound loci(Konopacki et al. 2019)34. This comparison aimed to identify FOXP3-dependent genes associated with H3K4me3 and H3K27me3 deposition. As expected, H3K4me3 was enriched at gene promoters (fig. S1, A and B). A Venn diagram revealed a significant overlap between FOXP3 binding sites and H3K4me3 peaks, with minimal overlap with H3K27me3 peaks (Fig. 1, A and B). The overlapping regions between FOXP3 binding sites and H3K4me3 or H3K27me3 peaks were predominantly located at promoters (Fig. 1, A and B). We then clustered the promoters into four groups based on the enrichment of H3K4me3 and H3K27me3. Clusters 1 and 3 showed strong enrichment of H3K4me3; cluster 2 was enriched with H3K27me3; and cluster 4 showed weak enrichment of both modifications. FOXP3 enrichment analysis revealed that FOXP3 preferentially binds to the promoters of clusters 1 and 3, which display high levels of H3K4me3 (Fig. 1C). Correspondingly, genes in these clusters exhibited high transcription levels, as demonstrated by our reanalysis of previously published RNA-seq data (Oh et al. 2017) shown in the right panels of Figure 1D35. In contrast, genes with H3K27me3 enrichment at their promoters were transcribed at low levels. Gene Ontology (GO) analysis of these four clusters revealed that cluster 1 was significantly enriched in genes involved in mRNA processing, covalent chromatin modification, and histone modification, while cluster 3 was enriched in genes related to DNA repair and mitochondrion organization (Fig. 1E). Cluster 2, on the other hand, was enriched in the pattern specification process, whereas cluster 4 showed no significant correlation with Treg cells. As expected, signature Treg cell genes such as Tnfrsf18, Nrp1, Stat5a, Lag3, Icos, and Pdcd1 were enriched in clusters 1 and 3, which exhibited strong H3K4me3 marks (fig. S1C). Conversely, FOXP3-bound sites in cluster 2 were specifically enriched for H3K27me3, including genes like Hic1, Trp73, and Rnf157, which are associated with inflammatory responses(fig. S1C). These observations support the hypothesis that FOXP3 primarily functions as an activator by promoting H3K4me3 deposition in Treg cells.

H3K4me3 is required for FOXP3-dependent gene activation in Treg cells.

(A)Venn diagram showing overlap of H3K4me3 enriched regions (this study) and FOXP3 binding sites (Konopacki et al.2019) in sorted CD4+YFP+ Treg cells (left). Genomic distribution of overlapped peaks(right). Note that the overlapped peaks are predominantly enriched at promoters.

(B)Venn diagram showing overlap of H3K27me3 enriched regions (Wei et al.2009) and FOXP3 binding sites in Treg cells (left). Genomic distribution of overlapped peaks (right). Note that the overlapped peaks are predominantly enriched at promoters.

(C)Heatmap showing enrichment of H3K27me3, H3K4me3, and FOXP3 surrounding the TSS. Unsupervised K-means clustering was conducted on H3K27me3 and H3K4me3 signals.

(D)Heatmap showing gene expression levels in Treg cells (RNA-seq data was obtained from Oh et al.2017) The clusters were consistent as in Fig. 1C.

(E)Gene Ontology (GO) pathway analysis of different clusters.

CXXC1 interacts with FOXP3 and binds H3K4me3-enriched sites in Treg cells

We conducted an enrichment analysis of known motifs at the overlapping peaks of FOXP3 ChIP-seq and H3K4me3 CUT&Tag in Treg cells to identify epigenetic factors that directly interact with FOXP3 to mediate chromatin remodeling and transcriptional reprogramming. Motif analysis of the overlapping peaks between FOXP3 binding sites and regions enriched in H3K4me3 revealed that, in addition to transcription factors, the most abundant motif associated with H3K4me3 was the epigenetic factor CXXC1(fig. S2A). To investigate this further, we performed CUT&Tag for endogenous CXXC1 in Treg cells to examine the genome-wide co-occupancy of CXXC1 and FOXP3. Over half of these CXXC1 binding sites were located at promoter regions (fig. S2B). Additionally, CXXC1 exhibited strong binding at transcription start site (TSS) and CpG islands (CGIs) (Fig. 2A and fig. S2C). As illustrated by the Venn diagram (Fig. 2B), more than half of the FOXP3-bound genes and H3K4me3-enriched genes were also bound by CXXC1. Similarly, more than half of CXXC1 peaks were overlapped with FOXP3 peaks (fig. S2D). Furthermore, the CXXC1-specific and FOXP3-specific binding sites also demonstrated modest binding of FOXP3 and CXXC1, respectively (Fig. 2C). These findings indicate that FOXP3 and CXXC1 share a substantial number of target genes in Treg cells. To confirm this interaction, we further validated the reciprocal immunoprecipitation of both endogenous and exogenous CXXC1 and FOXP3 (Fig. 2D and fig. S2E). An immunofluorescence assay revealed predominant colocalization of CXXC1 with FOXP3 in the nucleus (Fig. 2E). Overall, these results suggest that CXXC1 primarily functions as a coactivator of FOXP3-driven transcription in Treg cells.

CXXC1 interacts with FOXP3 in Treg cell.

(A)Average CXXC1 CUT&Tag signals around promoters in Treg cells. IgG was used as the control.

(B)Venn diagrams showing the overlap of FOXP3 binding genes, CXXC1 binding genes, and H3K4me3 enriched genes in Treg cells. Gene promoters covered by FOXP3 binding sites, CXXC1 binding sites, or exhibited high H3K4me3 levels were defined as FOXP3-bound genes, CXXC1-bound genes, or H3K4me3-enriched genes.

(C)Heatmaps showing FOXP3 ChIP-seq and CXXC1 CUT&Tag signals at indicated regions.

(D)Interaction between FOXP3 and CXXC1 was assessed by co-IP (forward and reverse) using Treg cell lysates.

(E)Immunofluorescence for FOXP3 and CXXC1 colocalization in Treg cells. Scale bars, 2 μm.

Complete ablation of Cxxc1 in Treg cells leads to a fatal autoimmune disease

To investigate the role of CXXC1 in Treg-cell homeostasis and function, we generated Foxp3CreCxxc1fl/fl mice (conditional knockout [cKO] mice) by crossing Cxxc1fl/fl with Foxp3YFP-Cre 36 mice, thereby specifically deleting Cxxc1 in Treg cells. The effective depletion of Cxxc1 in Treg cells was confirmed through quantitative PCR (qPCR) and Western blotting (fig. S3A). Notably, cKO mice appeared normal at birth but later exhibited spontaneous mortality starting around three weeks of age (Fig. 3A). Deletion of Cxxc1 in Treg cells led to the development of severe inflammatory disease, characterized by reduced body size, stooped posture, crusting of the eyelids, ears, and tail, and skin ulceration, particularly on the head and upper back (Fig. 3, B and C). Additionally, cKO mice developed extensive splenomegaly and lymphadenopathy (Fig. 3D). Histopathological analysis revealed massive lymphocyte and myeloid cell infiltration in the skin, lungs, liver sinusoids, and colon mucosa (Fig. 3E). Moreover, cKO mice displayed a significant increase in CD8+ T cell percentages (fig. S3B), along with a marked rise in cells exhibiting an effector/memory phenotype (CD44hi CD62Llo) (Fig. 3F). Furthermore, T cells from cKO mice produced elevated levels of IFN-γ, IL-17, and IL-4 in CD4+ T cells, as well as increased IFN-γ production in CD8+ T cells (Fig. 3G and fig. S3, C and D). These phenotypes closely resembled those observed in Foxp3-deficient mice 37 or mice with depleted Treg cells38, suggesting a deficiency in immune suppression.

Foxp3creCxxc1fl/fl mice spontaneously develop a fatal early-onset inflammatory disorder.

(A)Survival curves of Foxp3cre (black line) and Foxp3creCxxc1fl/fl (red line) mice (n = 20).

(B)Gross body weight of Foxp3cre and Foxp3creCxxc1fl/fl mice (n = 6).

(C)A representative image of Foxp3cre and Foxp3creCxxc1fl/fl mice.

(D)Representative images showing the spleen and peripheral lymph nodes from Foxp3cre and Foxp3creCxxc1fl/fl mice.

(E)Hematoxylin and eosin staining of the skin, lung, liver, and colon from Foxp3cre and Foxp3creCxxc1fl/fl mice (scale bar, 100 μm).

(F)Flow cytometry analysis of CD62L and CD44 expression on peripheral lymph node CD4+ and CD8+T cells from Foxp3cre and Foxp3creCxxc1fl/fl mice (left). Right, frequency of CD44lowCD62Lhi and CD44hiCD62Llow population in CD4+ or CD8+ T cells (n = 9).

(G)Lymph node cells from Foxp3cre and Foxp3creCxxc1fl/fl mice were stimulated ex vivo with PMA + ionomycin for 4h and analyzed for IFN-γ expressing in CD4+ or CD8+ T cells using flow cytometry (left). Right, percentages of IFN-γ+CD4+ or IFN-γ+CD8+ T cells in the lymph nodes of Foxp3cre and Foxp3creCxxc1fl/fl mice (n = 7).

All mice analyzed were 18-20 days old unless otherwise specified. Error bars show mean ± SD. The log-rank survival curve was used for survival analysis in A, and multiple unpaired t-test or two-tailed Student’s t-test were used for statistical analyses in B, F, and G (****P<0.0001). The flow cytometry results are representative of three independent experiments.

CXXC1 is necessary for the maintenance of Treg-cell suppressive activity

Despite the development of severe autoimmune disease, we observed an increase in both the absolute number and percentage of FOXP3+ Treg cells in the lymph nodes (fig. S4A). The expression level of the FOXP3 protein was only slightly altered in Cxxc1-deficient Treg cells (fig. S4B). In an in vitro suppression assay, Treg cells from Foxp3CreCxxc1fl/fl and WT mice exhibited similar suppressive effects on naïve T (Tn) cell proliferation (fig. S4C). The expression of the hallmark Treg-cell marker CTLA-4 showed a modest increase in Cxxc1-deficient Treg cells compared to WT Treg cells, while the expression of GITR remained unchanged (fig. S4D). To further assess the suppressive capacity of Cxxc1-deficient Treg cells in vivo, we employed the experimental autoimmune encephalomyelitis (EAE) model. Naïve CD4+ T cells from 2D2 mice were co-transferred with Treg cells from either Foxp3Cre or Foxp3Cre Cxxc1fl/fl mice into Rag1-/- recipients, and EAE was induced in these recipient mice. Mice that received only naïve CD4+ T cells from 2D2 mice developed more severe EAE symptoms (Fig. 4A). The addition of WT Treg cells from Foxp3Cre mice slightly mitigated EAE progression and reduced Th17 cells in the spinal cord (Fig. 4, A to D). In contrast, Foxp3Cre Cxxc1fl/fl Treg cells failed to suppress EAE (Fig. 4, A to D), and the cKO mice showed a significant reduction in Treg cells frequency in CNS tissues(Fig. 4E). Finally, we examined the role of CXXC1 in Treg cell-mediated suppression using T cell transfer-induced colitis, in which naïve T cells were transferred to Rag1-/- recipients either alone or together with WT or Foxp3Cre Cxxc1fl/fl Treg cells. The transfer of naïve T cells led to weight loss and intestinal pathology in recipient mice(Fig. 4, F and G). Mice receiving WT Treg cells continued to gain weight (Fig. 4F), whereas those that received Treg cells from cKO mice were unable to prevent colitis and exhibited a reduced percentage of Treg cells (Fig. 4, F to H). These findings underscore the critical role of CXXC1 in maintaining Treg cell function in vivo.

CXXC1 is essential for Treg cells to suppress T cell-mediated EAE and colitis.

(A)Mean clinical scores for EAE in Rag1−/− recipients of 2D2 CD4+ T cells, either alone or in combination with Foxp3cre or Foxp3creCxxc1fl/fl mice after immunization with MOG35–55, complete Freund’s adjuvant (CFA), and pertussis toxin (n=5).

(B and C)Representative histology of the spinal cord of Rag1−/− mice after EAE induction. Hematoxylin and eosin(H&E) staining (upper), Luxol fast blue (F&B) staining (lower). Scale bars, 50 μm (400×) and 200 μm (100×).

(D)Representative flow cytometry plots and quantification of the the percentages of IFNγ+ or IL-17A+ CD4+Vβ11+ T cells (n=5).

(E)Statistical analysis of the percentage CD4+ FOXP3+ Treg cell in CNS tissues 14 days after EAE induction.

(F)Changes in body weight of Rag1−/− mice after colitis induction(n=6).

(G)Haematoxylin and eosin (H&E) staining of colons from T cell-induced colitis mice 6 weeks after T cell transfer. Scale bars, 50 μm (400×) and 200 μm (100×).

(H)Statistical analysis of the percentage CD4+ FOXP3+ Treg cell in the spleen, mesenteric lymph nodes, and colon 6 weeks after colitis induction.

Error bars show mean ± SD. P values are determined by a two-tailed Student’s t-test or two-way ANOVA and Holm–Sidak post hoc test (A, D, and E). (*P<0.05, **P<0.01, ***P < 0.001, ****P<0.0001).

Treg cell lineage stability and proliferation depend upon CXXC1

Treg cells harbor a diverse T-cell receptor (TCR) repertoire, which likely play a critical role in their immune suppression function 39–41. To explore the role of CXXC1 in Treg-mediated suppression, we performed single-cell RNA sequencing (scRNA-seq) combined with TCR sequencing (TCR-seq) on CD4+YFP+ Treg cells isolated from mouse lymph nodes. After quality control and removal of doublets, 18,577 cells were retained for further analysis. Through unsupervised clustering and uniform manifold approximation and projection (UMAP) analysis, we identified eight distinct Treg cell clusters based on the expression of well-characterized markers, with a particular focus on two clusters of activated Treg cells that exhibited higher expression of markers and gene sets relative to naïve Treg cells (Fig. 5A and fig. S5, A to C). A comparison between Cxxc1-deficient and WT Treg cells within each cluster revealed a reduction in Cxxc1-deficient cells in the naïve subsets, while an increase was observed in the Gzmb+ and H2-Eb1+ subsets (Fig. 5B). To further elucidate the transition of Treg cells along a dynamic biological timeline, we constructed pseudo-time trajectories using Slingshot 42. The pseudo-time gradient depicted a progression from quiescent to activated Treg cells, ultimately encompassing the Gzmb+ and H2-Eb1+ subsets (Fig. 5C). Given the antigen-specific suppression capabilities of Treg cells 43,44, we examined their clonal expansion. The analysis revealed that expanded WT TCR clonotypes (n ≥ 2) were predominantly distributed among the Nt5e+ subsets, while Cxxc1-deficient Treg cells showed expanded clonotypes primarily within the Gzmb+ and H2-Eb1+ subsets(Fig. 5D and Fig. S5D). TCR sharing analysis indicated significant clonotype sharing among various clusters of WT Treg cells, suggesting a degree of homogeneity. However, the reduced TCR sharing in Cxxc1-deficient cells implies that decreased TCR diversity may impair the suppressive activity of Treg cells (Fig. 5E) 41. Furthermore, the Cxxc1-deficient group exhibited lower expression of several Treg-specific genes associated with suppressive functions, such asNt5e, Il10, Pdcd1, Klrg1, as well as genes that inhibit effector T cells differentiation, including Sell and Tcf7 (Fig. 5F and fig. S5E). Conversely, Cxxc1-deficient Treg cells demonstrated elevated expression of Gzmb, Il2ra, and Cd69 compared to WT, reflecting a profile indicative of increased activation (Fig. 5, F and G). Additionally, we observed increased expression of genes linked to Th1-type inflammation, such as Ifng, Tbx21, and Hif1a, in Cxxc1-deficient Treg cells, likely due to extreme inflammatory conditions (Fig. 5, F to H). The proportion of FOXP3+Ki67+ Treg cells was lower in cKO mice compared to WT mice (Fig. 5I). These findings underscore the crucial role of CXXC1 in maintaining Treg cell stability.

Single-cell transcriptomics reveals distinct Treg cell populations.

(A)UMAP plot showing clusters identified based on variable gene expression of sorted CD4+YFP+ Treg cells. Each dot represents a cell, and each color corresponds to a different population of cell types. Clustering analysis revealed 8 distinct Treg cell populations.

(B)Mean fold changes in cluster abundance between Foxp3cre and Foxp3creCxxc1fl/fl mice.

(C)Pseudotime trajectories of Treg cells based on Slingshot, color-coded by Treg cell subpopulations.

(D)Visualization of density and clonotype richness across Treg clusters from Foxp3cre and Foxp3creCxxc1fl/fl mice.

(E)TCR sharing of expanded clonotypes across all possible combinations of Treg cells from Foxp3cre and Foxp3creCxxc1fl/fl mice.

(F)Heatmap showing Z scores for the average expression of Treg-specific genes in each cluster between Foxp3cre and Foxp3creCxxc1fl/fl mice.

(G-H)Representative flow cytometry plots and quantification of (G) expression of CD25, CD69, ICOS, T-bet, and (H) IFN-γ in CD4+YFP+ Treg cells from Foxp3cre and Foxp3creCxxc1fl/fl mice (n=5 CD25, n=10 CD69, n=5 ICOS, n=6 T-bet, n=5 IFN-γ).

(I)Ki-67 expression (left) and frequency (right) in CD4+FOXP3+ Treg cells from Foxp3cre a-nd Foxp3creCxxc1fl/fl mice (n=6).

Error bars show mean ± SD. P values are determined by a two-tailed Student’s t-test (G-H). (**P<0.01, ***P < 0.001, ****P<0.0001). The flow cytometry results are representative of three independent experiments.

Cxxc1 deletion leads to a reduction in essential Treg cell markers

To confirm that the deficiency in Treg cell function in Treg-specific Cxxc1-deficient animals is due to intrinsic defects caused by Cxxc1 deficiency, rather than severe autoimmune inflammation in Foxp3Cre Cxxc1fl/fl mice, we examined Cxxc1-sufficient and Cxxc1-deficient Treg cell subsets in heterozygous Foxp3Cre/+ Cxxc1fl/fl (designated as “het-KO”) and littermate Foxp3Cre/+ Cxxc1fl/+ (designated as “het-WT”) female mice (Fig. 6A). Notably, het-KO female mice did not exhibit overt signs of autoimmunity, as random X-chromosome inactivation led to the coexistence of both Cxxc1-cKO and Cxxc1-WT Treg cells. However, both the frequency and absolute numbers of FOXP3+YFP+ Treg cells within the total Treg population were reduced in het-KO mice compared to their counterparts in het-WT littermates, indicating that Cxxc1 deficiency imposes a competitive disadvantage on Treg cells (Fig. 6B). Additionally, Cxxc1-deficient YFP+ Treg cells failed to upregulate the proliferation marker Ki-67 (Fig. 6C). Moreover, YFP+ Treg cells in het-KO female mice showed reduced expression of key genes essential for suppressive function, including Icos, Il2ra, Ctla4, and Tnfrsf18, compared to YFP-Treg cells from the same mice (Fig. 6D). To investigate the molecular program affected by the deletion of Cxxc1 in Treg cells, we performed RNA-sequencing (RNA-seq) analysis on CD4+YFP+ Treg cells isolated from het-WT and het-KO mice. We then conducted a differential gene expression (DGE) analysis based on the RNA-seq data. Among all expressed genes, 865 were upregulated (“Up” genes) and 761 were downregulated (“Down” genes) by ≥1.5-fold in CD4+YFP+ Treg cells from het-KO mice compared to het-WT mice, with a false discovery rate (FDR)–adjusted P-value cutoff of 0.05 (fig. S6A). Gene Ontology (GO) enrichment analysis revealed that the downregulated genes in Cxxc1-deficient Treg cells were predominantly enriched in pathways related to the negative regulation of immune system process and regulation of cell−cell adhesion (fig. S6B). The Cxxc1-deficient Treg cells also showed reduced expression of several genes associated with Treg cell suppressive function, including Il10, Tigit, Lag3, Icos, Nt5e(encoding CD73) and Itgae (encoding CD103) (fig. S6C). Thus, although YFP− WT Treg cells can prevent autoimmunity in het-KO mice, the absence of Cxxc1 in YFP+ Treg cells leads to the dysregulation of key Treg cell markers under steady-state conditions.

Cxxc1-deficient Treg cells in het-KO female mice exhibit reduced suppression markers under steady-state conditions.

(A)Schematic representation of wild-type and Cre-positive Treg cells in female Foxp3Cre/+mice.

(B)Flow cytometry analysis of the YFP-FOXP3+ (WT) and YFP+FOXP3+ (KO) Treg cells in Foxp3Cre/+ Cxxc1fl/+ (het-WT) and Foxp3Cre/+ Cxxc1fl/fl (het-KO) female mice (left), along with the frequency and absolute numbers of YFP+ cells within the total Treg population (right)(n=5).

(C)Flow cytometry analysis of Ki-67expression (left) and MFI (right) in YFP− and YFP+ cells within the CD4+FOXP3+ Treg cells from 6- to 8-week-old het-KO female mice (n=4).

(D)Representative flow cytometry plots and quantification of ICOS, CD25, CTLA4, and GITR expression in YFP− and YFP+ cells within CD4+FOXP3+ Treg cells from het-KO female mice (n=5).

Error bars show mean ± SD. P values are determined by a two-tailed Student’s t-test (G-H). (**P<0.01, ***P < 0.001, ****P<0.0001). The flow cytometry results are representative of three independent experiments.

The FOXP3-CXXC1 complex regulates the expression of key factors in Treg cells that are associated with the breadth of H3K4me3

CXXC1 binds to unmethylated CpG DNA via its N-terminal CXXC finger domain, facilitating its interaction with DNA methyltransferase 1 (DNMT1). This binding stabilizes the DNMT1 protein, thereby regulating DNA methylation 45,46. To investigate whether CXXC1 depletion affects DNA methylation in Treg cells, we performed whole genome bisulfite sequencing (WGBS) on Treg cells isolated from both WT and cKO mice. On average, Cxxc1-deficient Treg cells exhibited no significant changes in DNA methylation at gene loci or across genome-wide CpG sites, irrespective of chromosomal region (fig. S7, A to C). Furthermore, Cxxc1 knockout Treg cells did not show a significant increase in DNA methylation at key Treg signature gene loci (fig. S7D). Given the pivotal role of MLL4-mediated H3K4me1 in establishing the enhancer landscape and facilitating long-range chromatin interactions during Treg cell development15, we performed CUT&Tag to assess changes in H3K4me1 levels in Cxxc1-deficient Treg cells. This analysis revealed that H3K4me1 levels were similar in both WT and Cxxc1-deficient Treg cells (fig. S7, E to G).

While H3K4me3 modifications typically form sharp 1-to 2-kb peaks around promoters, some genes exhibit broader H3K4me3 regions, referred to as broad H3K4me3 domains (H3K4me3-BDs), which can extend to cover part or all of the gene’s coding sequences (up to 20 kb) 47,48. Broad H3K4me3 domains are preferentially associated with genes essential for the identity or function of specific cell types 47,49 and have been implicated in enhancing transcriptional elongation and increasing enhancer activity 49. To further explore the connection between broad H3K4me3 domains and the expression of immune-regulatory genes, we analyzed genes containing broad H3K4me3 regions. We classified the H3K4me3 domains surrounding transcription start sites (TSSs) into three categories: broad (more than 5 kb), medium (between 1 and 5 kb), and narrow (less than 1 kb) (Fig. 7A). Notably, Cxxc1-deficient Treg cells exhibited weaker H3K4me3 signals compared to WT cells within the broad H3K4me3 domains where CXXC1 binding is prominent (Fig. 7, A and B). Using the criteria established by Benayoun et al.47, which defines the top 5% of the widest H3K4me3 domains as BDs, we observed similar enrichment results (fig. S7, H and I). We then compared three groups of genes: BD-associated genes with reduced H3K4me3 levels following Cxxc1 deletion, genes with direct CXXC1 binding, and genes with direct FOXP3 binding. The Venn diagram revealed that the majority of genes (283 out of 294, 96%) with CXXC1 binding and reduced H3K4me3 levels overlap with FOXP3-bound genes, suggesting that CXXC1 presence enhances FOXP3 binding within the broad H3K4me3 domain (Fig. 7C). Furthermore, GO term analysis indicated that BD-associated genes are enriched in biological processes related to the negative regulation of immune system processes (Fig. 7D). Genome browser views displayed the enrichments of FOXP3, CXXC1, and H3K4me3 at key signature genes in Treg cells, such as Ctla4, Il2ra, Icos, and Tnfrsf18, with lower H3K4me3 densities observed at these loci in Cxxc1-deficient Treg cells (Fig. 7E). Similar patterns were observed at core genes involved in Treg homeostasis and suppressive function (e.g., Lag3, Nt5e, Ikzf4, and Cd28) (fig. S7, J)50,51. These findings suggest that CXXC1 and FOXP3 collaboratively promote sustained Treg cell homeostasis and function by preserving the H3K4me3 modification at key Treg cell genes.

CXXC1 regulates Foxp3-dependent molecule H3K4 trimethylation in Treg cells.

(A-B)Heatmaps showing H3K4me3(A)and CXXC1(B) signals centered on narrow, medium, and broad domains. The top panel shows the average CUT&Tag signals around indicated domains.

(C)Venn diagrams showing the overlap of FOXP3 binding genes, CXXC1 binding genes, and H3K4me3-BD-associated genes with decreased H3K4me3 levels after CXXC1 depletion in Treg cells.

(D)Gene Ontology (GO) pathway analysis of the overlap peaks.

(E)Representative genome browser view showing the enrichments of FOXP3, CXXC1, and H3K4me3 in Treg cells.

Discussion

Treg cells specifically express the transcription factor FOXP3, which is essential for maintaining Treg lineage stability and suppressive function52. However, FOXP3 alone is insufficient to fully regulate the transcriptional signature and functionality of Treg cells; its interaction with protein partners is crucial for this regulation. In this study, we identify CXXC1 as a key transcription factor that, by acting as a critical cofactor of FOXP3, orchestrates the Treg transcriptional program, which in turn plays a vital role in Treg cell homeostasis and function.

FOXP3 is known to promote histone H3 acetylation at the promoters and enhancers of its target genes, such as Il2ra, Ctla4, and Tnfrsf18, following Treg cell activation, thereby functioning as a transcriptional activator53,54. Conversely, FOXP3 can act as a repressor by silencing target genes like Il2 and Ifng through the induction of histone H3 deacetylation, mediated by the recruitment of histone deacetylases and transcriptional co-repressors10,55. Additionally, FOXP3 exerts this repression by recruiting the Ezh2-containing polycomb repressive complex to target genes during activation, as FOXP3-repressed genes are associated with H3K27me3 deposition and reduced chromatin accessibility11,56. Recent studies have begun to explore the biological significance of H3K4me3 breadth, revealing a positive correlation between H3K4me3 breadth and gene expression 49,57,58. These studies also suggest that H3K4me3 breadth contributes to defining specific cell identities during development and disease, including systemic autoimmune diseases like systemic lupus erythematosus and various cancers 47,59–62. Our study demonstrates that the loss of Cxxc1 leads to reduced H3K4me3 levels, predominantly in genes with broader peaks, such as Il2ra, Tnfrsf18, Ctla4 and Icos, directly impairing their suppressive function in Treg cells. Although our research primarily focused on the role of CXXC1 in Treg cells, it is plausible that similar mechanisms may be operative in other cell types.

The Treg-specific deletion of Cxxc1 leads to a rapid and fatal autoimmune disorder, characterized by systemic inflammation and tissue damage, underscoring the essential role of CXXC1 in maintaining immune self-tolerance within Treg cells. Interestingly, despite this severe phenotype, our findings show that H3K4me1 levels were comparable between WT and Cxxc1-deficient Treg cells. In contrast, MLL4 is critical for Treg cell development in the thymus, primarily by regulating H3K4me1, though it is not required for peripheral Treg cell function15. Numerous studies investigating the effectors of Treg cell-mediated suppression have identified a wide range of molecules and mechanisms. These include the upregulation of the inhibitory co-stimulatory receptor CTLA-4, which initiates inhibitory signaling; the sequestration of the T-cell growth factor IL-2 via CD25; the secretion of inhibitory cytokines, such as interleukin (IL)-10, IL-35 and TGF-β and the activity of ectoenzymes CD39 and CD73 on the Treg cells surface, which convert extracellular ATP, a potent pro-inflammatory mediator, into its anti-inflammatory counterpart, adenosine41,63. Mechanistically, we demonstrated that CXXC1 interacts with FOXP3 to regulate Treg cell function by trimethylating H3K4 at broad H3K4me3 domains of multiple genes involved in suppressive functions such as Il2ra, Nt5e, and Ctla4. Additionally, MLL1, another KMT, controls Treg cell activation and function by specifically regulating H3K4 trimethylation at genes encoding key Treg-related molecules such as Tigit, Klrg1, Tbx21, Cxcr3, and serves as a crucial epigenetic regulator in establishing a stable Th1-Treg lineage14.

Our results offer provide novel insights into the suppressive functions, heterogeneity, and regulatory mechanisms of Treg cells. Maintaining Treg cell homeostasis and function remains a significant challenge in harnessing Treg cells for the treatment of autoimmune diseases and the prevention of graft rejection.

eLife Assessment

This study presents important findings on the role of CXXC-finger protein 1 in regulatory T cell gene regulation and function. The evidence supporting the authors' claims is solid, with mostly state-of-the-art technology, although the inclusion of more mechanistic insights would have strengthened the study. The work will be of relevance to immunologists interested in regulatory T cell biology and autoimmunity.

Reviewer #1 (Public review):

Summary:

This work investigated the role of CXXC-finger protein 1 (CXXC1) in regulatory T cells. CXXC1-bound genomic regions largely overlap with Foxp3-bound regions and regions with H3K4me3 histone modifications in Treg cells. CXXC1 and Foxp3 interact with each other, as shown by co-immunoprecipitation. Mice with Treg-specific CXXC1 knockout (KO) succumb to lymphoproliferative diseases between 3 to 4 weeks of age, similar to Foxp3 KO mice. Although the immune suppression function of CXXC1 KO Treg is comparable to WT Treg in an in vitro assay, these KO Tregs failed to suppress autoimmune diseases such as EAE and colitis in Treg transfer models in vivo. This is partly due to the diminished survival of the KO Tregs after transfer. CXXC1 KO Tregs do not have an altered DNA methylation pattern; instead, they display weakened H3K4me3 modifications within the broad H3K4me3 domains, which contain a set of Treg signature genes. These results suggest that CXXC1 and Foxp3 collaborate to regulate Treg homeostasis and function by promoting Treg signature gene expression through maintaining H3K4me3 modification.

Strengths:

Epigenetic regulation of Treg cells has been a constantly evolving area of research. The current study revealed CXXC1 as a previously unidentified epigenetic regulator of Tregs. The strong phenotype of the knockout mouse supports the critical role CXXC1 plays in Treg cells. Mechanistically, the link between CXXC1 and the maintenance of broad H3K4me3 domains is also a novel finding.

Weaknesses:

(1) It is not clear why the authors chose to compare H3K4me3 and H3K27me3 enriched genomic regions. There are other histone modifications associated with transcription activation or repression. Please provide justification.

(2) It is not clear what separates Clusters 1 and 3 in Figure 1C. It seems they share the same features.

(3) The claim, "These observations support the hypothesis that FOXP3 primarily functions as an activator by promoting H3K4me3 deposition in Treg cells." (line 344), seems to be a bit of an overstatement. Foxp3 certainly can promote transcription in ways other than promoting H3K3me3 deposition, and it also can repress gene transcription without affecting H3K27me3 deposition. Therefore, it is not justified to claim that promoting H3K4me3 deposition is Foxp3's primary function.

(4) For the in vitro suppression assay in Figure S4C, and the Treg transfer EAE and colitis experiments in Figure 4, the Tregs should be isolated from Cxxc1 fl/fl x Foxp3 cre/wt female heterozygous mice instead of Cxxc1 fl/fl x Foxp3 cre/cre (or cre/Y) mice. Tregs from the homozygous KO mice are already activated by the lymphoproliferative environment and could have vastly different gene expression patterns and homeostatic features compared to resting Tregs. Therefore, it's not a fair comparison between these activated KO Tregs and resting WT Tregs.

(5) The manuscript didn't provide a potential mechanism for how CXXC1 strengthens broad H3K4me3-modified genomic regions. The authors should perform Foxp3 ChIP-seq or Cut-n-Taq with WT and Cxxc1 cKO Tregs to determine whether CXXC1 deletion changes Foxp3's binding pattern in Treg cells.

Reviewer #2 (Public review):

FOXP3 has been known to form diverse complexes with different transcription factors and enzymes responsible for epigenetic modifications, but how extracellular signals timely regulate FOXP3 complex dynamics remains to be fully understood. Histone H3K4 tri-methylation (H3K4me3) and CXXC finger protein 1 (CXXC1), which is required to regulate H3K4me3, also remain to be fully investigated in Treg cells. Here, Meng et al. performed a comprehensive analysis of H3K4me3 CUT&Tag assay on Treg cells and a comparison of the dataset with the FOXP3 ChIP-seq dataset revealed that FOXP3 could facilitate the regulation of target genes by promoting H3K4me3 deposition.

Moreover, CXXC1-FOXP3 interaction is required for this regulation. They found that specific knockdown of Cxxc1 in Treg leads to spontaneous severe multi-organ inflammation in mice and that Cxxc1-deficient Treg exhibits enhanced activation and impaired suppression activity. In addition, they have also found that CXXC1 shares several binding sites with FOXP3 especially on Treg signature gene loci, which are necessary for maintaining homeostasis and identity of Treg cells.

The findings of the current study are pretty intriguing, and it would be great if the authors could fully address the following comments to support these interesting findings.

Major points:

(1) There is insufficient evidence in the first part of the Results to support the conclusion that "FOXP3 functions as an activator by promoting H3K4Me3 deposition in Treg cells". The authors should compare the results for H3K4Me3 in FOXP3-negative conventional T cells to demonstrate that at these promoter loci, FOXP3 promotes H3K4Me3 deposition.

(2) In Figure 3 F&G, the activation status and IFNγ production should be analyzed in Treg cells and Tconv cells separately rather than in total CD4+ T cells. Moreover, are there changes in autoantibodies and IgG and IgE levels in the serum of cKO mice?

(3) Why did Cxxc1-deficient Treg cells not show impaired suppression than WT Treg during in vitro suppression assay, despite the reduced expression of Treg cell suppression assay -associated markers at the transcriptional level demonstrated in both scRNA-seq and bulk RNA-seq?

(4) Is there a disease in which Cxxc1 is expressed at low levels or absent in Treg cells? Is the same immunodeficiency phenotype present in patients as in mice?

Reviewer #3 (Public review):

In the report entitled "CXXC-finger protein 1 associates with FOXP3 to stabilize homeostasis and suppressive functions of regulatory T cells", the authors demonstrated that Cxxc1-deletion in Treg cells leads to the development of severe inflammatory disease with impaired suppressive function. Mechanistically, CXXC1 interacts with Foxp3 and regulates the expression of key Treg signature genes by modulating H3K4me3 deposition. Their findings are interesting and significant. However, there are several concerns regarding their analysis and conclusions.

Major concerns:

(1) Despite cKO mice showing an increase in Treg cells in the lymph nodes and Cxxc1-deficient Treg cells having normal suppressive function, the majority of cKO mice died within a month. What causes cKO mice to die from severe inflammation?

Considering the results of Figures 4 and 5, a decrease in Treg cell population due to their reduced proliferative capacity may be one of the causes. It would be informative to analyze the population of tissue Treg cells.

(2) In Figure 5B, scRNA-seq analysis indicated that Mki67+ Treg subset are comparable between WT and Cxxc1-deficient Treg cells. On the other hand, FACS analysis demonstrated that Cxxc1-deficient Treg shows less Ki-67 expression compared to WT in Figure 5I. The authors should explain this discrepancy.

In addition, the authors concluded on line 441 that CXXC1 plays a crucial role in maintaining Treg cell stability. However, there appears to be no data on Treg stability. Which data represent the Treg stability?

(3) The authors found that Cxxc1-deficient Treg cells exhibit weaker H3K4me3 signals compared to WT in Figure 7. This result suggests that Cxxc1 regulates H3K4me3 modification via H3K4 methyltransferases in Treg cells. The authors should clarify which H3K4 methyltransferases contribute to the modulation of H3K4me3 deposition by Cxxc1 in Treg cells.

Furthermore, it would be important to investigate whether Cxxc1-deletion alters Foxp3 binding to target genes.

(4) In Figure 7, the authors concluded that CXXC1 promotes Treg cell homeostasis and function by preserving the H3K4me3 modification since Cxxc1-deficient Treg cells show lower H3K4me3 densities at the key Treg signature genes. Are these Cxxc1-deficient Treg cells derived from mosaic mice? If Cxxc1-deficient Treg cells are derived from cKO mice, the gene expression and H3K4me3 modification status are inconsistent because scRNA-seq analysis indicated that expression of these Treg signature genes was increased in Cxxc1-deficient Treg cells compared to WT (Figure 5F and G).

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

This work investigated the role of CXXC-finger protein 1 (CXXC1) in regulatory T cells. CXXC1-bound genomic regions largely overlap with Foxp3-bound regions and regions with H3K4me3 histone modifications in Treg cells. CXXC1 and Foxp3 interact with each other, as shown by co-immunoprecipitation. Mice with Treg-specific CXXC1 knockout (KO) succumb to lymphoproliferative diseases between 3 to 4 weeks of age, similar to Foxp3 KO mice. Although the immune suppression function of CXXC1 KO Treg is comparable to WT Treg in an in vitro assay, these KO Tregs failed to suppress autoimmune diseases such as EAE and colitis in Treg transfer models in vivo. This is partly due to the diminished survival of the KO Tregs after transfer. CXXC1 KO Tregs do not have an altered DNA methylation pattern; instead, they display weakened H3K4me3 modifications within the broad H3K4me3 domains, which contain a set of Treg signature genes. These results suggest that CXXC1 and Foxp3 collaborate to regulate Treg homeostasis and function by promoting Treg signature gene expression through maintaining H3K4me3 modification.

Strengths:

Epigenetic regulation of Treg cells has been a constantly evolving area of research. The current study revealed CXXC1 as a previously unidentified epigenetic regulator of Tregs. The strong phenotype of the knockout mouse supports the critical role CXXC1 plays in Treg cells. Mechanistically, the link between CXXC1 and the maintenance of broad H3K4me3 domains is also a novel finding.

Weaknesses:

(1) It is not clear why the authors chose to compare H3K4me3 and H3K27me3 enriched genomic regions. There are other histone modifications associated with transcription activation or repression. Please provide justification.

Thank you for highlighting this important point. We chose to focus on H3K4me3 and H3K27me3 enriched genomic regions because these histone modifications are well-characterized markers of transcriptional activation and repression, respectively. H3K4me3 is predominantly associated with active promoters, while H3K27me3 marks repressed chromatin states, particularly in the context of gene regulation at promoters. This duality provides a robust framework for investigating the balance between transcriptional activation and repression in Treg cells. While histone acetylation, such as H3K27ac, is linked to enhancer activity and transcriptional elongation, our focus was on promoter-level regulation, where H3K4me3 and H3K27me3 are most relevant. Although other histone modifications could provide additional insights, we chose to focus on these two to maintain clarity and feasibility in our analysis. We have revised the text accordingly; please refer to Page 18, lines 353-356.

(2) It is not clear what separates Clusters 1 and 3 in Figure 1C. It seems they share the same features.

We apologize for not clarifying these clusters clearly. Cluster 1 and 3 are both H3K4me3 only group, with H3K4me3 enrichment and gene expression levels being higher in Cluster 1. At first, we divided the promoters into four categories because we wanted to try to classify them into four categories: H3K4me3 only, H3K27me3 only, H3K4me3-H3K27me3 co-occupied, and None. However, in actual classification, we could not distinguish H3K4me3-H3K27me3 co-occupied group. Instead, we had two categories of H3K4me3 only, with cluster 1 having a higher enrichment level for H3K4me3 and gene expression levels.

(3) The claim, "These observations support the hypothesis that FOXP3 primarily functions as an activator by promoting H3K4me3 deposition in Treg cells." (line 344), seems to be a bit of an overstatement. Foxp3 certainly can promote transcription in ways other than promoting H3K3me3 deposition, and it also can repress gene transcription without affecting H3K27me3 deposition. Therefore, it is not justified to claim that promoting H3K4me3 deposition is Foxp3's primary function.

Thank you for your insightful feedback. We agree that the statement in line 344 may have overstated the role of FOXP3 in promoting H3K4me3 deposition as its primary function. As you pointed out, FOXP3 is indeed a multifaceted transcription factor that regulates gene expression through various mechanisms. It can promote transcription independent of H3K4me3 deposition, as well as repress transcription without directly influencing H3K27me3 levels.

To more accurately reflect the broader regulatory functions of FOXP3, we have revised the manuscript. The updated text (Page 19, lines 385-388) now reads:

"These findings collectively support the conclusion that FOXP3 contributes to transcriptional activation in Treg cells by promoting H3K4me3 deposition at target loci, while also regulating gene expression directly or indirectly through other epigenetic modifications.

(4) For the in vitro suppression assay in Figure S4C, and the Treg transfer EAE and colitis experiments in Figure 4, the Tregs should be isolated from Cxxc1 fl/fl x Foxp3 cre/wt female heterozygous mice instead of Cxxc1 fl/fl x Foxp3 cre/cre (or cre/Y) mice. Tregs from the homozygous KO mice are already activated by the lymphoproliferative environment and could have vastly different gene expression patterns and homeostatic features compared to resting Tregs. Therefore, it's not a fair comparison between these activated KO Tregs and resting WT Tregs.

Thank you for raising this insightful point regarding the potential activation status of Treg cells in homozygous knockout mice. To address this concern, we performed additional experiments using Treg cells isolated from Foxp3Cre/+Cxxc1fl/fl (hereafter referred to as “het-KO”) female mice and their littermate controls, Foxp3Cre/+Cxxc1fl/+ (referred to as “het-WT”) mice.

The results of these new experiments are now included in the manuscript (Page25, lines 507–509, Figure 6E and Figure S6A-E):

(1) In the in vitro suppression assay, Treg cells from het-KO mice exhibited reduced suppressive function compared to het-WT Treg cells. This finding underscores the intrinsic defect in Treg cells suppressive capacity attributable to the loss of one Cxxc1 allele.

(2) In the experimental autoimmune encephalomyelitis (EAE) model, Treg cells isolated from het-KO mice also demonstrated impaired suppressive function.

(5) The manuscript didn't provide a potential mechanism for how CXXC1 strengthens broad H3K4me3-modified genomic regions. The authors should perform Foxp3 ChIP-seq or Cut-n-Taq with WT and Cxxc1 cKO Tregs to determine whether CXXC1 deletion changes Foxp3's binding pattern in Treg cells.

Thank you for raising this important point. To address your suggestion, we performed CUT&Tag experiments and found that Cxxc1 deletion does not alter FOXP3 binding patterns in Treg cells. Most FOXP3-bound regions in WT Treg cells were similarly enriched in KO Treg cells, indicating that Cxxc1 deficiency does not impair FOXP3’s DNA-binding ability. These results have been added to the revised manuscript (Page 28, lines 567-575, Figure S8A-B) and are further discussed in the Discussion (Pages 28-29, lines 581-587).

Reviewer #2 (Public review):

FOXP3 has been known to form diverse complexes with different transcription factors and enzymes responsible for epigenetic modifications, but how extracellular signals timely regulate FOXP3 complex dynamics remains to be fully understood. Histone H3K4 tri-methylation (H3K4me3) and CXXC finger protein 1 (CXXC1), which is required to regulate H3K4me3, also remain to be fully investigated in Treg cells. Here, Meng et al. performed a comprehensive analysis of H3K4me3 CUT&Tag assay on Treg cells and a comparison of the dataset with the FOXP3 ChIP-seq dataset revealed that FOXP3 could facilitate the regulation of target genes by promoting H3K4me3 deposition.

Moreover, CXXC1-FOXP3 interaction is required for this regulation. They found that specific knockdown of Cxxc1 in Treg leads to spontaneous severe multi-organ inflammation in mice and that Cxxc1-deficient Treg exhibits enhanced activation and impaired suppression activity. In addition, they have also found that CXXC1 shares several binding sites with FOXP3 especially on Treg signature gene loci, which are necessary for maintaining homeostasis and identity of Treg cells.

The findings of the current study are pretty intriguing, and it would be great if the authors could fully address the following comments to support these interesting findings.

Major points:

(1) There is insufficient evidence in the first part of the Results to support the conclusion that "FOXP3 functions as an activator by promoting H3K4Me3 deposition in Treg cells". The authors should compare the results for H3K4Me3 in FOXP3-negative conventional T cells to demonstrate that at these promoter loci, FOXP3 promotes H3K4Me3 deposition.

Thank you for this insightful comment. We have already performed additional experiments comparing H3K4Me3 levels between FOXP3-positive Treg cells and FOXP3-negative conventional T cells (Tconv). Please refer to Pages 18, lines 361-368, and Figure 1C and Figure S1C for the results. Our results show that H3K4Me3 abundance is higher at many Treg-specific gene loci in Treg cells compared to Tconv cells. This supports our conclusion that FOXP3 promotes H3K4Me3 deposition at these loci.

(2) In Figure 3 F&G, the activation status and IFNγ production should be analyzed in Treg cells and Tconv cells separately rather than in total CD4+ T cells. Moreover, are there changes in autoantibodies and IgG and IgE levels in the serum of cKO mice?

Thank you for your valuable suggestions. In response to your comment, we reanalyzed the data in Figures 3F and 3G to assess the activation status and IFN-γ production in Tconv cells. The updated analysis revealed that Cxxc1 deletion in Treg cells leads to increased activation and IFN-γ production in Tconv cells. Additionally, we corrected the analysis of IL-17A and IL-4 expression, which were upregulated in Tconv cells. These updated results are now included in the revised manuscript (Page 21, lines 429-431, Figure 3I and Figure S3E-F).

Additionally, we examined autoantibodies and immunoglobulin levels in the serum of Cxxc1 cKO mice. Our data show a significant increase in serum IgG levels, accompanied by elevated IgG autoantibodies, indicating heightened autoimmune responses. In contrast, serum IgE levels remained largely unchanged. The results are detailed in the revised manuscript (Page 21, lines 421-423, Figure 3E and Figure S3B).

(3) Why did Cxxc1-deficient Treg cells not show impaired suppression than WT Treg during in vitro suppression assay, despite the reduced expression of Treg cell suppression assay -associated markers at the transcriptional level demonstrated in both scRNA-seq and bulk RNA-seq?

Thank you for your thoughtful comment. The absence of impaired suppression in Cxxc1-deficient Treg cells from homozygous knockout (KO) mice during the in vitro suppression assay, despite the reduced expression of Treg-associated markers at the transcriptional level (as demonstrated by scRNA-seq), can likely be explained by the activated state of these Treg cells. In homozygous KO mice, Treg cells are already activated due to the lymphoproliferative environment, resulting in gene expression patterns that differ from those of resting Treg cells. This pre-activation may obscure the effect of Cxxc1 deletion on their suppressive function in vitro.

To address this limitation, we used heterozygous Foxp3Cre/+Cxxc1fl/fl (het-KO) female mice, along with their littermate controls, Foxp3Cre/+Cxxc1fl/+ (het-WT) mice. In these heterozygous mice, we observed an impairment in Treg cell suppressive function in vitro, which was accompanied by the downregulation of several key Treg-associated genes, as confirmed by RNA-Seq analysis.

These updated findings, based on the use of het-KO mice, are now incorporated into the revised manuscript (Page 25, lines 507–509, Figure 6E).

(4) Is there a disease in which Cxxc1 is expressed at low levels or absent in Treg cells? Is the same immunodeficiency phenotype present in patients as in mice?

This is indeed a very meaningful and intriguing question, and we are equally interested in understanding whether low or absent Cxxc1 expression in Treg cells is associated with any human diseases. However, despite an extensive review of the literature and available data, we found no reports linking Cxxc1 deficiency in Treg cells to immunodeficiency phenotypes in patients comparable to those observed in mice.

Reviewer #3 (Public review):

In the report entitled "CXXC-finger protein 1 associates with FOXP3 to stabilize homeostasis and suppressive functions of regulatory T cells", the authors demonstrated that Cxxc1-deletion in Treg cells leads to the development of severe inflammatory disease with impaired suppressive function. Mechanistically, CXXC1 interacts with Foxp3 and regulates the expression of key Treg signature genes by modulating H3K4me3 deposition. Their findings are interesting and significant. However, there are several concerns regarding their analysis and conclusions.

Major concerns:

(1) Despite cKO mice showing an increase in Treg cells in the lymph nodes and Cxxc1-deficient Treg cells having normal suppressive function, the majority of cKO mice died within a month. What causes cKO mice to die from severe inflammation?

Considering the results of Figures 4 and 5, a decrease in the Treg cell population due to their reduced proliferative capacity may be one of the causes. It would be informative to analyze the population of tissue Treg cells.

Thank you for your insightful observation regarding the mortality of cKO mice despite increased Treg cells in lymph nodes and the normal suppressive function of Cxxc1-deficient Treg cells.

As suggested, we hypothesized that the reduction of tissue-resident Treg cells could be a key factor. Additional experiments revealed a significant decrease in Treg cell populations in the small intestine lamina propria (LPL), liver, and lung of cKO mice. These findings highlight the critical role of tissue-resident Treg cells in preventing systemic inflammation.

This reduction aligns with Figures 4 and 5, which demonstrate impaired proliferation and survival of Cxxc1-deficient Treg cells. Together, these defects lead to insufficient Treg populations in peripheral tissues, escalating localized inflammation into systemic immune dysregulation and early mortality.

These additional results have been incorporated into the revised manuscript (Page21, lines 424-427, Figure 3G and Figure S3C).

(2) In Figure 5B, scRNA-seq analysis indicated that the Mki67+ Treg subset is comparable between WT and Cxxc1-deficient Treg cells. On the other hand, FACS analysis demonstrated that Cxxc1-deficient Treg shows less Ki-67 expression compared to WT in Figure 5I. The authors should explain this discrepancy.

Thank you for pointing out the apparent discrepancy between the scRNA-seq and FACS analyses regarding Ki-67 expression in Cxxc1-deficient Treg cells.

In Figure 5B, the scRNA-seq analysis identified the Mki67+ Treg subset as comparable between WT and Cxxc1-deficient Treg cells. This finding reflects the overall proportion of cells expressing Mki67 transcripts within the Treg population. In contrast, the FACS analysis in Figure 5I specifically measures Ki-67 protein levels, revealing reduced expression in Cxxc1-deficient Treg cells compared to WT.

To resolve this discrepancy, we performed additional analyses of the scRNA-seq data to directly compare the expression levels of Mki67 mRNA between WT and Cxxc1-deficient Treg cells. The results revealed a consistent reduction in Mki67 transcript levels in Cxxc1-deficient Treg cells, aligning with the reduced Ki-67 protein levels observed by FACS.

These new analyses have been included in the revised manuscript (Author response image 1) to clarify this point and demonstrate consistency between the scRNA-seq and FACS data.

Violin plots displaying the expression levels of Mki67 in Treg cells from Foxp3cre and Foxp3creCxxc1fl/fl mice.

In addition, the authors concluded on line 441 that CXXC1 plays a crucial role in maintaining Treg cell stability. However, there appears to be no data on Treg stability. Which data represent the Treg stability?

Thank you for your valuable comment. We agree that our wording in line 441 may have been too conclusive. Our data focus on the impact of Cxxc1 deficiency on Treg cell homeostasis and transcriptional regulation, rather than directly measuring Treg cell stability. Specifically, the downregulation of Treg-specific suppressive genes and upregulation of pro-inflammatory markers suggest a shift in Treg cell function, which points to disrupted homeostasis rather than stability.

We have revised the manuscript to clarify that CXXC1 plays a crucial role in maintaining Treg cell function and homeostasis, rather than stability (Page 24, lines 489-491).

(3) The authors found that Cxxc1-deficient Treg cells exhibit weaker H3K4me3 signals compared to WT in Figure 7. This result suggests that Cxxc1 regulates H3K4me3 modification via H3K4 methyltransferases in Treg cells. The authors should clarify which H3K4 methyltransferases contribute to the modulation of H3K4me3 deposition by Cxxc1 in Treg cells.

We appreciate the reviewer’s insightful comment regarding the role of H3K4 methyltransferases in regulating H3K4me3 deposition by CXXC1 in Treg cells.

CXXC1 has been reported to function as a non-catalytic component of the Set1/COMPASS complex, which includes the H3K4 methyltransferases SETD1A and SETD1B—key enzymes responsible for H3K4 trimethylation(1-4). Based on these findings, we propose that CXXC1 modulates H3K4me3 levels in Treg cells by interacting with and stabilizing the activity of the Set1/COMPASS complex.

These revisions are further discussed in the Discussion (Page 30-31, lines 624-632).

Furthermore, it would be important to investigate whether Cxxc1-deletion alters Foxp3 binding to target genes.

Thank you for raising this important point. To address your suggestion, we performed CUT&Tag experiments and found that Cxxc1 deletion does not alter FOXP3 binding patterns in Treg cells. Most FOXP3-bound regions in WT Treg cells were similarly enriched in KO Treg cells, indicating that Cxxc1 deficiency does not impair FOXP3’s DNA-binding ability. These results have been added to the revised manuscript (Page 28, lines 567-575, Figure S8A-B) and are further discussed in the Discussion (Pages 28-29, lines 581-587).

(4) In Figure 7, the authors concluded that CXXC1 promotes Treg cell homeostasis and function by preserving the H3K4me3 modification since Cxxc1-deficient Treg cells show lower H3K4me3 densities at the key Treg signature genes. Are these Cxxc1-deficient Treg cells derived from mosaic mice? If Cxxc1-deficient Treg cells are derived from cKO mice, the gene expression and H3K4me3 modification status are inconsistent because scRNA-seq analysis indicated that expression of these Treg signature genes was increased in Cxxc1-deficient Treg cells compared to WT (Figure 5F and G).

Thank you for your insightful comment. To clarify, the Cxxc1-deficient Treg cells analyzed for H3K4me3 modifications in Figure 7 were derived from Cxxc1 conditional knockout (cKO) mice, not mosaic mice.

Regarding the apparent inconsistency between reduced H3K4me3 levels and the increased expression of Treg signature genes observed in scRNA-seq analysis (Figure 5F and G), we believe this discrepancy can be attributed to distinct mechanisms regulating gene expression. H3K4me3 is an epigenetic mark that facilitates chromatin accessibility and transcriptional regulation, reflecting upstream chromatin dynamics. However, gene expression levels are influenced by a combination of factors, including transcriptional activators, downstream compensatory mechanisms, and the inflammatory environment in cKO mice.

The upregulation of Treg signature genes in scRNA-seq data likely reflects an activated or pro-inflammatory state of Cxxc1-deficient Treg cells in response to systemic inflammation, as previously described in the manuscript. This contrasts with the intrinsic reduction in H3K4me3 levels at these loci, indicating a loss of epigenetic regulation by CXXC1.

To further support this interpretation, RNA-seq analysis of Treg cells from Foxp3Cre/+ Cxxc1fl/fl (“het-KO”) and their littermate Foxp3Cre/+ Cxxc1fl/+ (“het-WT”) female mice (Figure S6C) revealed a significant reduction in key Treg signature genes such as Icos, Ctla4, Tnfrsf18, and Nt5e in het-KO Treg cells. These results align with the diminished H3K4me3 modifications observed in cKO Treg cells, further underscoring the role of CXXC1 as an epigenetic regulator.

In summary, while the gene expression changes observed in scRNA-seq may reflect adaptive responses to inflammation, the reduced H3K4me3 modifications directly highlight the critical role of CXXC1 in maintaining the epigenetic landscape essential for Treg cell homeostasis and function.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

In Figure 7E, the y-axis scale for H3K4me3 peaks at the Ctla4 locus should be consistent between WT and cKO samples.

We thank the reviewer for pointing out the inconsistency in the y-axis scale for the H3K4me3 peaks at the Ctla4 locus in Figure 7E. We have carefully revised the figure to ensure that the y-axis scale is now consistent between the WT and cKO samples.

We appreciate the reviewer’s attention to this detail, as it enhances the rigor of the data presentation. Please find the updated Figure 7E in the revised manuscript.

Reviewer #2 (Recommendations for the authors):

In lines 455 and 466, the name of Treg signature markers validated by flow cytometry should be written as protein name and capitalized.

Thank you for pointing this out. We have carefully reviewed lines 455 and 466 and have revised the text to ensure that the Treg signature markers validated by flow cytometry are referred to using their protein names, with proper capitalization.

Reviewer #3 (Recommendations for the authors):

(1) On line 431, "Cxxc1-deficient cells" should be Cxxc1-deficient Treg cells".

We thank the reviewer for highlighting this oversight. On line 431, we have revised "Cxxc1-deficient cells" to "Cxxc1-deficient Treg cells" to provide a more accurate and specific description. We appreciate the reviewer's attention to detail, as this correction improves the precision of our manuscript.

(2) In Figure 4H, negative values should be removed from the y-axis.

Thank you for your observation. We have revised Figure 4H to remove the negative values from the y-axis, as requested. This adjustment ensures a more accurate and meaningful representation of the data.

(3) It is better to provide the lists of overlapping genes in Figure 7C.

Thank you for your suggestion. We agree that providing the lists of overlapping genes in Figure 7C would enhance the clarity and reproducibility of the results. We have now included the gene lists as supplementary information (Supplementary Table 3) accompanying Figure 7C.

(1) Lee, J. H. & Skalnik, D. G. CpG-binding protein (CXXC finger protein 1) is a component of the mammalian set1 histone H3-Lys4 methyltransferase complex, the analogue of the yeast Set1/COMPASS complex. Journal of Biological Chemistry 280, 41725-41731, doi:10.1074/jbc.M508312200 (2005).

(2) Thomson, J. P., Skene, P. J., Selfridge, J., Clouaire, T., Guy, J., Webb, S., Kerr, A. R. W., Deaton, A., Andrews, R., James, K. D., Turner, D. J., Illingworth, R. & Bird, A. CpG islands influence chromatin structure via the CpG-binding protein Cfp1. Nature 464, 1082-U1162, doi:10.1038/nature08924 (2010).

(3) Shilatifard, A. in Annual Review of Biochemistry, Vol 81 Vol. 81 Annual Review of Biochemistry (ed R. D. Kornberg) 65-95 (2012).

(4) Brown, D. A., Di Cerbo, V., Feldmann, A., Ahn, J., Ito, S., Blackledge, N. P., Nakayama, M., McClellan, M., Dimitrova, E., Turberfield, A. H., Long, H. K., King, H. W., Kriaucionis, S., Schermelleh, L., Kutateladze, T. G., Koseki, H. & Klose, R. J. The SET1 Complex Selects Actively Transcribed Target Genes via Multivalent Interaction with CpG Island Chromatin. Cell Reports 20, 2313-2327, doi:10.1016/j.celrep.2017.08.030 (2017).

Version 2

Abstract

Abstract

FOXP3-expressing regulatory T (Treg) cells play a pivotal role in maintaining immune homeostasis and tolerance, with their activation being crucial for preventing various inflammatory responses. However, the mechanisms governing the epigenetic program in Treg cells during their dynamic activation remain unclear. In this study, we demonstrate that CXXC finger protein 1 (CXXC1) interacts with the transcription factor FOXP3 and facilitates the regulation of target genes by modulating H3K4me3 deposition. Cxxc1 deletion in Treg cells leads to severe inflammatory disease and spontaneous T-cell activation, with impaired immunosuppressive function. As a transcriptional regulator, CXXC1 promotes the expression of key Treg functional markers under steady-state conditions, which are essential for the maintenance of Treg cell homeostasis and their suppressive functions. Epigenetically, CXXC1 binds to the genomic regulatory regions of Treg program genes in mouse Treg cells, overlapping with FOXP3 binding sites. Given its critical role in Treg cell homeostasis, CXXC1 presents itself as a promising therapeutic target for autoimmune diseases.

Introduction

Regulatory T (Treg) cells are a distinct subset of CD4+ T cells that play a critical role in maintaining immune homeostasis and self-tolerance by suppressing excessive or aberrant immune responses to foreign or self-antigens 1–3. These cells can be further categorized into thymus-derived regulatory T cells (tTreg cells), periphery-derived Treg cells (pTreg cells), and induced Treg cells (iTreg cells) 4. They uniquely express the transcription factor FOXP3, a member of the forkhead winged-helix family, which is essential for Treg cell lineage commitment and suppressive function 5,6. Deletion or mutation of the Foxp3 gene leads to a range of immunological disorders, including allergies, immunopathology, and autoimmune diseases in both mice and humans 7,8.

It is well established that FOXP3 recruits various cofactors to form complexes that either promote or repress the expression of downstream genes, with histone and DNA modifications playing pivotal roles in this process. FOXP3 can activate or repress the transcription of key regulators of Treg cell activation and function by recruiting the histone acetyltransferases (HATs) or histone deacetylases (HDACs)9,10. Notably, FOXP3-bound sites exhibit enrichment of H3K27me3, a modification essential for FOXP3-mediated repressive chromatin remodeling under inflammatory conditions11. However, the direct role of FOXP3 as a transcriptional activator through interactions with epigenetic regulators, particularly via modulation of H3K4 trimethylation, remains poorly documented.

In mammals, six proteins have been identified that catalyze H3K4 methylation. These proteins contain the SET domain and include MLL1 (KMT2A), MLL2 (KMT2B), MLL3 (KMT2C), MLL4 (KMT2D), SETD1A, and SETD1B 12,13. For example, MLL1 plays a critical role as an epigenetic regulator in Treg cell activation and functional specialization14. Additionally, Placek et al. demonstrated that MLL4 is essential for Treg cell development by catalyzing H3K4me1 at distant unbound enhancers through chromatin looping 15. H3K4me3, which is enriched at the transcription start site (TSS) and the CpG island (CGI), converts chromatin into active euchromatin by recruiting activating factors 16. CXXC finger protein 1(CXXC1, also known as CFP1), which contains a SET1 interaction domain (SID), is required for binding to the histone H3K4 methyltransferases SETD1A and SETD1B 17. Previous studies have shown that CXXC1 regulates promoter patterns during T cell maturation, GM-CSF-derived macrophage phagocytosis, TH17 cell differentiation, and the function of ILC3 cells with aging by modulating H3K4me3 modification18–21. Despite the well-documented role of CXXC1 in various immune effector cells, its role in Treg cells remains unclear.

Here, we demonstrate that CXXC1 interacts with FOXP3 and enhances the expression of FOXP3 target genes. Treg cell-specific deletion of Cxxc1 triggers systemic autoimmunity, accompanied by multiorgan inflammation, ultimately resulting in early-onset fatal inflammatory disease in mice. Cxxc1-deficient Treg cells exhibit a disadvantage in proliferation and homeostasis, even in non-inflammatory mice where coexisting wild-type (WT) Treg cells were present. Moreover, Cxxc1-deficient Treg cells display intrinsic defects in the expression of key suppression molecules, including CTLA-4, CD25, ICOS, and GITR. Consistent with these findings, mechanistic studies strongly suggest that CXXC1 functions as an essential cofactor of FOXP3, playing a crucial role in sustaining H3K4me3 modifications at key Treg cell genes.

Materials and Methods

Mice

All mice used in this study were bred for a minimum of seven generations on a C57BL/6 background. Mouse experiments, or cells from mice of the same genotype, compared littermates or age-matched control animals. The Cxxc1fl/fl mouse strain has been previously described 22. The Foxp3cre mice (JAX,016959) were generously provided by Bin Li (Shanghai Jiao Tong University School of Medicine, Shanghai, China). CD45.1 (NM-KI-210226) mice were purchased from the Nanjing Biomedical Research Institute of Nanjing University. Rag1−/-mice (stock#T004753) were purchased from GemPharmatech. 2D2 (MOG35-55-specific TCR transgenic) mice were graciously supplied by Prof. Linrong Lu (Zhejiang University School of Medicine, Hangzhou, Zhejiang, China). Foxp3cre (WT) and Foxp3cre Cxxc1fl/fl (cKO) were utilized at 3 weeks old unless otherwise stated. All mice were housed in the Zhejiang University Laboratory Animal Center under specific pathogen-free conditions, and all animal experimental procedures were approved by the Zhejiang University Animal Care and Use Committee (approval no.ZJU20230246).

Cell culture

HEK 293T cells were obtained from ATCC, and Plat E cells were kindly provided by Prof. Xiaolong Liu (Shanghai Institutes for Biological Sciences). Both cell lines were cultured in Dulbecco-modified Eagle medium (DMEM) containing 10% (v/v) fetal bovine serum (FBS), supplemented with 1% penicillin/streptomycin.

Immunofluorescence (IF) Staining

As previously described 23, coverslips were treated with a 0.01% poly-L-lysine solution (P4707; Sigma) for 10 minutes, air-dried, and then coated with CD4+ YFP+ Treg cells. The cells were then fixed in 4% formaldehyde for 15 minutes at room temperature, permeabilized with 0.2% Triton X-100, and blocked with 1% BSA. Antibodies against CXXC1 (ab198977; Abcam) and FOXP3( 17-5773-82; Invitrogen) were diluted in Image iT FX signal enhancer (I3693; Invitrogen) and incubated with cells overnight at 4 °C. After washing with phosphate-buffered saline (PBS), the cells were incubated with a goat anti-rabbit antibody Alexa Fluor 594 (1:250;10015289; Invitrogen) secondary antibody and stained with DAPI (200 ng/mL; D523; Dojindo). Slides were washed with PBS and sealed with an antifade solution(P36934; Invitrogen) before imaging with an Olympus FV3000 fluorescence microscope. The images were visualized using the FV31-SW software.

Co-immunoprecipitation and western blot

Harvest the appropriate transfected cell lines and primary cells from culture and wash them with ice-cold PBS. Lyse the cells in NETN300 buffer (300CmM NaCl, 0.5CmM EDTA, 0.5% (v/v) NP-40, 20CmM Tris-HCl pH 8.0) supplemented with a protease inhibitor (1:100, P8340, Sigma-Aldrich) and PMSF (1 mM) on ice for 10 minutes. Take a small portion of the whole-cell lysate as input, and incubate the remaining lysate with either Anti-FLAG M2 Beads (M8823; Sigma) or Anti-HA Beads (HY-K0201; MCE) on a rotator at 4°C overnight. For the endogenous Co-IP assay targeting CXXC1 and FOXP3, incubate the cell lysate with protein G magnetic beads along with anti-CXXC1 (ab198977; Abcam) or anti-FOXP3 (14-4774-82; Invitrogen) antibodies on a rotator at 4°C overnight. Wash the beads three times with IP buffer (100CmM NaCl, 0.5CmM EDTA, 0.5% (v/v) NP-40, 20CmM Tris-HCl pH 8.0) to remove non-specific binding. Boil the washed beads with 1× Laemmli sample buffer (1610747; Bio-Rad) to elute the bound proteins. Separate the denatured proteins by SDS-PAGE. Transfer the separated proteins onto PVDF membranes (Millipore) for immunoblotting.Immunoblot the PVDF membranes (IPVH00010) with the following antibodies: anti-CXXC1 (1:1000; ab198977; Abcam), anti-FOXP3 (1:500; 14-7979-80; Invitrogen), anti-FlAG (1:1000; 14793; Cell Signaling Technology), anti-HA (1:1000; 3724S; Cell Signaling Technology).Detect the immunoblotted proteins using a secondary HRP-conjugated goat anti-rabbit antibody (1:1000; HA1001-100; Huabio) and visualize the bands using an appropriate detection method.

ELISA

Serum samples from 3-week-old WT and KO mice were analyzed for total IgG and IgE concentrations using ELISA kits (88-50630-88; eBioscience) according to the manufacturer’s instructions. Half-area ELISA plates were coated with Coating Buffer and incubated overnight at 4°C. After washing with PBST (PBS, 1 mM EDTA, 0.05% Tween-20), the plates were blocked with 5% BSA in PBS for 30 minutes at room temperature. Serum was diluted to the appropriate concentration with blocking buffer and incubated overnight at 4°C. After washing with PBST, the plates were incubated with HRP-conjugated anti-mouse IgG(1033-05; SouthernBiotech) and IgE(1110-05; SouthernBiotech) antibodies (1:2000 in 1% BSA/PBST) at 37°C for 1 hour. Following washing, TMB substrate was added, and the reaction was stopped with 2 M HCSOC after sufficient color development (1–15 minutes). Absorbance at 450 nm was measured within 30 minutes.

Lymphocyte Isolation and Flow Cytometry

Cells from lymphoid organs were prepared by mechanical disruption between frosted slides, while non-lymphoid organs were processed enzymatically. For lung tissue, minced samples were digested in RPMI containing 100 μg/mL DNase I (9003-98-9;Sigma-Aldrich) and 2 mg/mL Collagenase D (LS004188;Worthington Biochemical) at 37°C for 1.5 hours. Liver tissue was minced and digested in RPMI supplemented with 100 μg/mL DNase I and 1 mg/mL Collagenase D at 37°C for 30 minutes, with lymphocytes isolated using a 40-70% Percoll (GE Healthcare) gradient. For intestinal tissue, the samples were first incubated in DMEM containing 3% FBS, 0.2% HEPES, 0.5 M EDTA, and 0.145 mg/mL dithiothreitol for 10 minutes. This was followed by digestion with 50 mg/mL DNase I and 145 mg/mL Collagenase II (Worthington Biochemical) in DMEM at 37°C for 5 minutes. Lymphocytes were then isolated using an 80% and 40% Percoll gradient.

For surface marker analysis, cells were incubated for 15 minutes with purified anti-mouse CD16/32 antibody (101320; BioLegend) to block Fc receptors. After blocking, cells were stained with the indicated antibodies for surface markers. To determine cytokine expression, cells were stimulated for 4 hours at 37C with phorbol12-myristate13-acetate (50ng/mL; S1819; Beyotime), ionomycin (1mg/ml; S1672; Beyotime), and brefeldin A (BFA;00-4506-51; Invitrogen). After stimulation, cells were labeled with a fixable viability dye and surface markers. Cells were then fixed and permeabilized according to the manufacturer’s instructions (00-8222-49; Invitrogen). For transcription factor staining, samples were fixed using the FOXP3/Transcription Factor Staining Buffer Set (00-5523-00; Invitrogen). Flow cytometry was conducted using a BD Fortessa (BD Biosciences) or LongCyte (Beijing Challen Biotechnology Co., Ltd.) system. Flow cytometry data were acquired and analyzed using FlowJo software.

The following antibodies were purchased from Invitrogen or BioLegend: Zombie Viole t f-ixable viability(423113), Zombie NIR fixable(423105), CD25(PC61), CD8α(53-6.7), CD62L(MEL-14), PD-1(J43), CD44(IM7), IL-17A(TC1118H10), KL-RG1(2F1), CD4 (GK1.5), TCRβ(H57-597), IFN-γ(XMG1.2), FOXP3(FJK-16s), CTLA-4(UC10-4B9), IC OS(15F9), GITR(DTA-1), LAG3(C9B7W), TCRVβ11(RR3-15), CD45RB(C363-16A) C D69 (H1.2F3), MHCII(M5/114.15.2), CD80(16-10A1), T-bet(eBio4B10), IKZF4(ESB7C 2), IL-4(11B11), CCR7(4B12), CD73(TY/11.8).

Real-time PCR

Total RNA was extracted from Treg cells using the RNAAiso Plus(9109; Takara) reagent according to the manufacturer’s instructions, and cDNA synthesis was performed using the Prime Script RT Reagent Kit (Takara). TB Green Premix Ex Taq (RR420A; Takara) was used for quantitative real-time PCR (qPCR). The expression levels of target mRNA were normalized to the level of β-actin expression. The primers for qPCR are as follows:

Cxxc1 qPCR Forward: ATCCGGGAATGGTACTGTCG

Cxxc1 qPCR Reverse: CTGTGGAGAAGATTTGTGGG

β-actin qPCR Forward: CTGTCCCTGTATGCCTCTG

β-actin qPCR Reverse: ATGTCACGCACGATTTCC

CD4+T and YFP+ Treg cells adoptive transfer in Experimental autoimmune encephalomyelitis (EAE)

CD4+YFP+ Treg cells from Foxp3cre and Foxp3cre Cxxc1fl/fl mice were enriched using the Mouse CD4 T Cell Isolation Kit (480005; Biolegend) and then sorted using the BD Aria II flow cytometer. Naïve CD4+T cells from 2D2 (MOG35-55-specificTCR transgenic) mice were isolated using the Mouse CD4 Naïve T cell Isolation Kit480039 (480039; Biolegend). As previously described24, 2D2 naïve CD4+ T cells alone (5 × 105 per mouse), or 2D2 naïve CD4+ T cells (5× 105 per mouse) together with WT or Foxp3creCxxc1fl/fl Treg cells (2×105 per mouse), were transferred into Rag1−/− mice via the tail vein. One day after cell transfer, the recipient mice were inoculated subcutaneously (s.c.) with 200μg MOG35-55 peptide (MEVGWYRSPFSRVVHLYRNGK; GenemeSynthesis) emulsified in complete Freund’s adjuvant (CFA) (F5506; Sigma). Intravenous administration of 200 ng of Pertussis toxin (181; List Biological Laboratories) was performed on days 0 and 2 after peptide inoculation. The severity of EAE was monitored and blindly graded using a clinical score from 0 to 5: 0, no clinical signs; 1, limp tail; 2, paraparesis (weakness, incomplete paralysis of one or two hind limbs); 3, paraplegia (complete paralysis of two hind limbs); 4, paraplegia with forelimb weakness or paralysis; 5, dying or death.

Isolation lymphocytes from the central nervous system (CNS)

On day 14 after EAE induction, mice were perfused with transcardially administered PBS to eliminate contaminating blood cells in the central nervous system (CNS). The forebrain and cerebellum were dissected to expose the spinal cord, which was then carefully removed from the spinal canal. The fresh spinal cord was harvested and cut into 2 mm pieces. The CNS tissue pieces were homogenized using a syringe and passed through a 70 µM cell strainer to obtain a single-cell suspension. The single-cell suspension was digested with collagenase D (2μg/ml;11088858001; Roche) and deoxyribonuclease I (DNase I; 1 μg/ml; DN25; Sigma-Aldrich) at 37°C for 20 minutes under rotation. After digestion, the cell suspension was centrifuged to pellet the cells. The cell pellets were resuspended in 40% Percoll and layered onto a discontinuous Percoll gradient. Centrifugation at 80% Percoll allowed for the separation of cells at the 40–80% Percoll interface, which were collected as CNS mononuclear cells. The collected CNS mononuclear cells were washed with PBS to remove any remaining Percoll. CNS mononuclear cells were stimulated for 4 hours with PMA and ionomycin in the presence of Brefeldin A to induce cytokine production. After stimulation, cells were fixed, rendered permeable, and stained with appropriate antibodies for intracellular cytokine detection.

Histological analyses

The lungs, skin, liver, and colon were excised from three-week-old mice. Prior to histological analysis, the samples were fixed in formalin, embedded in paraffin, and stained with hematoxylin and eosin. For CNS histology, spinal cords were fixed in 4% paraformaldehyde, paraffin-embedded, sectioned, and stained with Luxol Fast Blue and hematoxylin and eosin (H&E). To examine colon histology, colons from Rag1−/− hosts were similarly processed and stained with H&E.

Adoptive transfer colitis model

Colitis was induced following the protocol described25. In brief, CD4+ YFP+ Treg cells were isolated from 3-week-old CD45.2+Foxp3creCxxc1fl/fl and CD45.2+ Foxp3cre mice. A total of 2×105 Treg cells from each group were mixed with 4×105 Teff cells (CD45.1+CD4+CD45RBhi) sorted from CD45.1+ mice and transferred into the Rag1−/- mice via intraperitoneal injection. Teff cells alone were transferred as a control group. Mouse body weight was measured weekly post-adoptive transfer. The percentage change in body weight was calculated by comparing the current weight with the initial weight on day 0. Mice were euthanized when any had reached 80% of their initial body weight. The large intestines were sectioned into 4μm thick slices and stained with hematoxylin.

In vitro Treg suppression assay

Naïve CD4+ T cells isolated from WT mice were labeled with CFSE (C34554; Invitrogen). CD4+YFP+ Treg cells from Foxp3cre and Foxp3creCxxc1fl/fl mice were cultured with naïve CD4+ T cells (1×105 cells) at various ratios in the presence of 2μg/mL anti-CD3(16-0031-85; Invitrogen) and 3μg/mL anti-CD28(16-0281-85; Invitrogen). On day 3, cells were analyzed by flow cytometry.

CUT&Tag

CUT&Tag assays of CD4+YFP+ Treg cells were conducted as previously described26. Briefly, approximately 1×105 single cells were carefully pipetted into wash buffer twice. The pelleted cells were resuspended in wash buffer, activated concanavalin (BP531; Bangs Laboratories), and incubated for 15 minutes at room temperature. Cells bound to the beads were resuspended in Dig-Wash Buffer and incubated with a 1:50 dilution of primary antibodies (rabbit anti-H3K4me3, Active Motif,39016; rabbit anti-CXXC1, abcam, ab198977; rabbit anti-FOXP3, abcam, ab 150743;normal IgG, Cell Signaling, 2729) at 4 C overnight. The beads were incubated with a secondary antibody (goat anti-rabbit IgG; SAB3700883; Sigma-Aldrish) diluted 1:100 in Dig-Wash buffer for 60 minutes at room temperature. Cells were treated with Hyperactive pG-Tn5 Transposase (S602; Vazyme) diluted in Dig-300 Buffer for 1 hour at room temperature. The cells were subsequently resuspended in Tagmentation buffer (10mM MgCl2 in Dig-300 Buffer) and incubated at 37C for 1 hour. To halt tagmentation, 10μl was spiked with 0.5M EDTA, 3μl with 10%SDS, and 3μl with 20mg/ml Proteinase K and incubated at 55C for 1h. DNA library amplification was performed according to the manufacturer’s instructions and purified using VAHTS DNA Clean Beads (N411; Vazyme). Libraries were sequenced on the Illumina NovaSeq platform ( Annoroad Gene Technology).

CUT&Tag and ChIP-seq data analysis

FOXP3 ChIP-seq data was obtained from GSE121279. H3K27me3 ChIP-seq data was obtained from GSE14254. CUT&Tag and ChIP-seq reads were trimmed to 50 bp and aligned against the mouse genome build mm9 using Bowtie2(v2.3.4.1) with default parameters. All PCR duplicates and unmapped reads were removed. Peak calling was performed using MACS2(v2.1.1.20160309) and signal tracks for each sample were generated using the ‘wigToBigWig’ utility of UCSC. We classified the H3k4me3 peaks around TSSs into three groups: broad (>5 kb), medium (1-5kb), and narrow (<1 kb). The top 5% of the widest peaks were considered as broad peaks. The average intensity profiles were generated using deepTools (v2.5.4). Motif analysis was performed using the ‘findmotifsGenome.pl’ command inHomer2 package. Epigenetic factors were identified using the Epigenetic Factor Database(https://epifactors.autosome.org/) and then screened for those that exclusively regulate the expression of their target genes by modulating the deposition of H3K4me3. Genomic distribution was analyzed using the “genomation” R package. GO pathway analysis was performed using the “clusterProfiler” R package. The sequencing information of CUT&Tag data used in this study is summarized in Supplementary Table S1.

Clustering analysis

Promoters were defined as ±2 kb regions flanking the annotated TSS. Reads in promoters were counted using the “coverage” command in bedtools (v2.26.0) and further normalized to RPKM. The k-means clustering of H3K4me3 and H3K27me3 enrichment at promoters was conducted using the “kmeans” function in R.

RNA-seq and data analysis

Total RNA was extracted from sorted CD4CYFPC Treg cells using the RNeasy Plus Mini Kit (Qiagen, #74134), following the manufacturer’s protocol. RNA-Seq libraries were constructed and sequenced by Haplox (Nanchang, China), using an Illumina platform with paired-end reads of 150 bp. RNA-seq data was obtained from GSE82076. Raw reads were trimmed to 50 bp and mapped to the mouse genome (mm9) using TopHat (v2.1.1) with default parameters. Only uniquely mapped reads were kept for downstream analysis. The RNA abundance of each gene was quantified using Cufflinks (v2.2.1).

Single-Cell RNA-Sequencing

A total of 300,000 sort-purified CD4+YFP+ Treg cells from Foxp3cre and Foxp3cre Cxxc1fl/fl were resuspended in BD Pharmingen Stain Buffer (FBS) (554656; BD). Single cells were isolated using a chromium controller (BD platform, BD Bioscience) according to the manufacturer’s instructions, as previously described27. The single cells were labeled with sample tags using the BD Mouse Immune Single-Cell Multiplexing Kit (633793; BD). Following standard protocols, cDNA amplification and library construction were performed to generate scRNA-seq libraries.

Targeted scRNA-seq data processing

The raw FASTQ files were processed by BD Rhapsody using the Targeted analysis pipeline. After alignment and filtering, the distribution-based error correction(DBEC)-adjusted molecules were loaded into R studio(version 4.3.2). All subsequent analyses were performed using the package Seurat (version 4.4.0) with default parameters. Specifically, the scRNA-seq data counts were log-normalized. All targeted genes were scaled and then were used for Principal components analysis (PCA). The batch effects were removed by the HarmonyMatrix function in the Harmony package (version 1.1.0). The first 20 principal components were used to calculate nonlinear dimensionality reduction using RunUMAP. Differential gene expression (DEGs) between clusters was assessed using the FindAllMarkers function. The clusters were then annotated based on DEGs. Barplots were generated using ggplot2 (version 3.4.4). Heatmaps were generated using pheatmap (version 1.0.12).

Analysis of the single-cell TCR-seq repertoire

Raw V(D)J fastq reads were processed using BD Rhapsody Pipeline and then were analyzed using scRepertoire (version 1.12.0). The TCR clonotype was called using the nucleotide sequence of the CDR3 region for both TCR alpha and beta chains. For cells with multiple chains, the top two clonotypes with the highest expression were selected for downstream analysis. Clonal overlap between different cell types was calculated using the clonalOverlap function of scRepertoire. A clonotype was defined as expansion if it could be detected in at least two cells.

scRNA-seq trajectory analysis

UMAP embeddings obtained from the Seurat package were projected into the Slingshot(version 2.10.0) package to construct pseudotime Trajectories for Treg cells. Naïve subsets were set as the root state.

Whole genome bisulfite sequencing (WGBS) and data analysis

Sorted CD4+YFP+ Treg cells (3×106) were lysed in cell lysis buffer to release DNA. The bisulfite-treated DNA was used to prepare the sequencing library. DNA libraries were transferred to the Illumina Platform for sequencing using 150 bp paired-end reads. Raw reads were trimmed using TrimGalore (v0.4.4) with default parameters. Subsequently, the reads were mapped against the mm9 reference genome using Bismark v0.19.0 with parameters ‘--bowtie2’. PCR duplicates were removed and the methylation levels were calculated using ‘bismark_methylation_extractor’. We calculated the mean CpG methylation levels of various genome elements: promoter, 5′-UTR, exon, intron, 3′-UTR, genebody, intergenic, CGIs, and repeats using in-house scripts. The sequencing information of WGBS data used in this study is summarized in Supplementary Table S2.

Statistical analysis

The statistical significance analysis was performed using Prism 8.0 (GraphPad). Error bars are presented as mean ± SD. P values of < 0.05 were deemed statistically significant (*P < 0.05, **P < 0.01, ***P < 0.001 and ****P < 0.000). Statistical analyses were performed with unpaired Student’s t-test or two-way ANOVA and Holm–Sidak post hoc test.

Results

FOXP3 binds regulatory loci primed for activation and repression in Treg cells

FOXP3-mediated gene expression is well-recognized, with several studies highlighting its dual role as both a transcriptional activator and repressor 28–32. However, the mechanisms by which FOXP3 regulates Treg-specific gene transcription via epigenetic modifications remain incompletely understood. To investigate these mechanisms, we employed CUT&Tag to generate genome-wide H3K4me3 maps in Treg cells. To complement this, we compared our data with an H3K27me3 ChIP-Seq dataset from Wei et al.33 focusing on previously identified FOXP3-bound loci(Konopacki et al. 2019)34. This integrative analysis allowed us to identify FOXP3-dependent genes associated with either H3K4me3 (indicative of transcriptional activation) or H3K27me3 (indicative of repression) deposition. These findings provide insight into how FOXP3 modulates the epigenetic landscape to regulate Treg cell function. As expected, H3K4me3 was enriched at gene promoters (fig. S1, A and B). A Venn diagram revealed overlap between FOXP3 binding sites and H3K4me3 peaks, with minimal overlap with H3K27me3 peaks (Fig. 1, A and B). The overlapping regions between FOXP3 binding sites and H3K4me3 or H3K27me3 peaks were predominantly located at promoters (Fig. 1, A and B). To further investigate the role of FOXP3 in promoting H3K4me3 deposition, we compared H3K4me3 levels between FOXP3-positive Treg cells and FOXP3-negative conventional T cells (Tconv). Consistent with our hypothesis, H3K4me3 abundance was higher at Treg-specific gene loci (e.g., Tnfrsf18, Ctla4, Il2ra and Nt5e) in Treg cells compared to Tconv cells (Fig. 1,C and fig. S1, C). Notably, the selection of these loci was guided by prior studies identifying genes specifically associated with Treg cell function35. This finding highlights the role of FOXP3 in shaping the epigenetic landscape of Treg cells through enhanced H3K4me3 deposition at these critical loci.

H3K4me3 is required for FOXP3-dependent gene activation in Treg cells.

(A) Venn diagram showing overlap of H3K4me3 enriched regions (this study) and FOXP3 binding sites (Konopacki et al.2019) in sorted CD4+YFP+Treg cells (left). Genomic distribution of overlapped peaks(right). Note that the overlapped peaks are predominantly enriched at promoters.

(B) Venn diagram showing overlap of H3K27me3 enriched regions (Wei et al.2009) and FOXP3 binding sites in Treg cells (left). Genomic distribution of overlapped peaks (right). Note that the overlapped peaks are predominantly enriched at promoters.

(C) Representative genome browser view showing the enrichments of H3K4me3 and FOXP3 inTconv or Treg cells.

(D) Heatmap showing enrichment of H3K27me3, H3K4me3, and FOXP3 surrounding the TSS. Unsupervised K-means clustering was conducted on H3K27me3 and H3K4me3 signals. (E)Heatmap showing gene expression levels in Treg cells (RNA-seq data was obtained from Oh et al.2017) The clusters were consistent as in Fig. 1C.

(E) Gene Ontology (GO) pathway analysis of different clusters.

To further characterize these modifications,we clustered promoters into four groups based on the enrichment of H3K4me3 and H3K27me3. Clusters 1 and 3 showed strong enrichment of H3K4me3; cluster 2 was enriched with H3K27me3; and cluster 4 showed weak enrichment of both modifications. FOXP3 preferentially bound to clusters 1 and 3 prom-oters, which displayed high H3K4me3 levels (Fig. 1C). Correspondingly, genes in these clusters exhibited high transcription levels, as shown by reanalysis of previously published RNA-seq data (Oh et al. 2017) 36 (Figure 1D, right). In contrast, genes with H3K27me3 enrichment at their promoters were transcribed at low levels.

Gene Ontology (GO) analysis of these four clusters revealed distinct functional roles. Cluster 1 was enriched in genes involved in mRNA processing, covalent chromatin modification, and histone modification, while cluster 3 was enriched in genes related to DNA repair and mitochondrion organization (Fig. 1E). Cluster 2, enriched with H3K27me3, was associated with pattern specification process, while cluster 4 showed no correlation with Treg cells. Notably, signature Treg cell genes such as Tnfrsf18, Nrp1, Stat5a, Lag3, Icos, and Pdcd1 were enriched in clusters 1 and 3, showing strong H3K4me3 marks (fig. S1C). Conversely, genes like Hic1, Trp73, and Rnf157, associated with inflammatory responses, were enriched for H3K27me3 in clusters 2 (fig. S1C). These findings collectively support the conclusion that FOXP3 contributes to transcriptional activation in Treg cells by promoting H3K4me3 deposition at target loci, while also regulating gene expression directly or indirectly through other epigenetic modifications.

CXXC1 interacts with FOXP3 and binds H3K4me3-enriched sites in Treg cells

We conducted an enrichment analysis of known motifs at the overlapping peaks of FOXP3 ChIP-seq and H3K4me3 CUT&Tag in Treg cells to identify epigenetic factors that directly interact with FOXP3 to mediate chromatin remodeling and transcriptional reprogramming. Motif analysis of the overlapping peaks between FOXP3 binding sites and regions enriched in H3K4me3 revealed that, in addition to transcription factors, the most abundant motif associated with H3K4me3 was the epigenetic factor CXXC1(fig. S2A). To investigate this further, we performed CUT&Tag for endogenous CXXC1 in Treg cells to examine the genome-wide co-occupancy of CXXC1 and FOXP3. Over half of these CXXC1 binding sites were located at promoter regions (fig. S2B). Additionally, CXXC1 exhibited strong binding at transcription start site (TSS) and CpG islands (CGIs)( Fig. 2A and fig. S2C). As illustrated by the Venn diagram (Fig. 2B), more than half of the FOXP3-bound genes and H3K4me3-enriched genes were also bound by CXXC1. Similarly, more than half of CXXC1 peaks were overlapped with FOXP3 peaks (fig. S2D). Furthermore, the CXXC1-specific and FOXP3-specific binding sites also demonstrated modest binding of FOXP3 and CXXC1, respectively (Fig. 2C). These findings indicate that FOXP3 and CXXC1 share a substantial number of target genes in Treg cells. To confirm this interaction, we further validated the reciprocal immunoprecipitation of both endogenous and exogenous CXXC1 and FOXP3 (Fig. 2D and fig. S2E). An immunofluorescence assay revealed predominant colocalization of CXXC1 with FOXP3 in the nucleus (Fig. 2E). Overall, these results suggest that CXXC1 primarily functions as a coactivator of FOXP3-driven transcription in Treg cells.

CXXC1 interacts with FOXP3 in Treg cell.

(A) Average CXXC1 CUT&Tag signals around promoters in Tregcells. IgG was used as the control.

(B) Venn diagrams showing the overlap of FOXP3 binding genes, CXXC1 binding genes, and H3K4me3 enriched genes in Treg cells. Gene promoters covered by FOXP3 binding sites, CXXC1 binding sites, or exhibited high H3K4me3 levels were defined as FOXP3-bound genes, CXXC1-bound genes, or H3K4me3-enriched genes.

(C) Heatmaps showing FOXP3 ChIP-seq and CXXC1 CUT&Tag signals at indicated regions.

(D) Interaction between FOXP3 and CXXC1 was assessed by co-IP (forward and reverse) using Treg cell lysates.

(E) Immunofluorescence for FOXP3 and CXXC1 colocalization in Treg cells. Scale bars, 2 μm.

Complete ablation of Cxxc1 in Treg cells leads to a fatal autoimmune disease

To investigate the role of CXXC1 in Treg-cell homeostasis and function, we generated Foxp3CreCxxc1fl/fl mice (conditional knockout [cKO] mice) by crossing Cxxc1fl/fl with Foxp3YFP-Cre 37 mice, thereby specifically deleting Cxxc1 in Treg cells. The effective depletion of Cxxc1 in Treg cells was confirmed through quantitative PCR (qPCR) and Western blotting (fig. S3A). Notably, cKO mice appeared normal at birth but later exhibited spontaneous mortality starting around three weeks of age (Fig. 3A). Deletion of Cxxc1 in Treg cells led to the development of severe inflammatory disease, characterized by reduced body size, stooped posture, crusting of the eyelids, ears, and tail, and skin ulceration, particularly on the head and upper back (Fig. 3, B and C). Additionally, cKO mice developed extensive splenomegaly and lymphadenopathy (Fig. 3D). Foxp3CreCxxc1fl/fl mice exhibited elevated serum levels of anti-dsDNA autoantibodies and IgG , along with a modest increase in IgE concentration (Figure 3E and fig. S3B). Histopathological analysis revealed massive lymphocyte and myeloid cell infiltration in the skin, lungs, liver sinusoids, and colon mucosa (Fig. 3F). In full agreement with the aforementioned severe autoimmune diseases, Foxp3CreCxxc1fl/fl mice had decreased percentages and numbers of CD4+ Foxp3+ Treg cells in small intestine lamina propria (LPL), liver and lung(Fig. 3G and fig. S3C). Moreover, cKO mice displayed an increase in CD8+ T cell percentages (fig. S3D), along with a marked rise in cells exhibiting an effector/memory phenotype (CD44hi CD62Llo) (Fig. 3H). Furthermore, T cells from cKO mice produced elevated levels of IFN-γ, IL-17, and IL-4 in CD4+ YFP− T cells, as well as increased IFN-γ production in CD8+ T cells (Fig. 3I and fig. S3, E and F). These phenotypes closely resembled those observed in Foxp3-deficient mice 38 or mice with depleted Treg cells39, suggesting a deficiency in immune suppression.

Foxp3creCxxc1fl/fl mice spontaneously develop a fatal early-onset inflammatory disorder.

(A) Survival curves of Foxp3cre (black line) and Foxp3creCxxc1fl/fl (red line) mice (n = 20).

(B) Gross body weight of Foxp3cre and Foxp3creCxxc1fl/fl mice (nC=C6).

(C) A representative image of Foxp3cre and Foxp3creCxxc1fl/fl mice.

(D) Representative images showing the spleen and peripheral lymph nodes from Foxp3cre and Foxp3creCxxc1fl/fl mice.

(E) ELISA quantification of anti-dsDNA IgG in the serum of Foxp3cre and Foxp3creCxxc1fl/fl mice (n = 4).

(F) Hematoxylin and eosin staining of the skin, lung, liver, and colon from Foxp3cre an d Foxp3creCxxc1fl/fl mice (scale bar, 100Cμm).

(G) Representative flow cytometry plots of CD4C Foxp3C Treg cells isolated from the small intestinal lamina propria (LPL), liver, and lung of Foxp3cre and Foxp3creCxxc1fl/fl mice.

(H) Flow cytometry analysis of CD62L and CD44 expression on peripheral lymph node CD4+YFP− and CD8+T cells from Foxp3cre and Foxp3creCxxc1fl/fl mice (left). Right, frequency of CD44lowCD62Lhi and CD44hiCD62Llow population in CD4+ YFP− or CD8+ T cells (n = 6).

(I) Lymph node cells from Foxp3cre and Foxp3creCxxc1fl/fl mice were stimulated ex vivo with PMA + ionomycin for 4h and analyzed for IFN-γ expressing in CD4+ YFP− or CD8+ T cells using flow cytometry (left). Right, percentages of IFN-γ+CD4+ YFP− or IFN-γ+CD8+ T cells in the lymph nodes of Foxp3cre and Foxp3creCxxc1fl/fl mice (n = 6).

All mice analyzed were 18-20 days old unless otherwise specified. Error bars show mean ± SD. The log-rank survival curve was used for survival analysis in A, and multiple unpaired t-test or two-tailed Student’s t-test were used for statistical analyses in B, E, Gand H (**P<0.01, ***PC<C0.001, ****P<0.0001). The flow cytometry results are representative of three independent experiments.

CXXC1 is necessary for the maintenance of Treg-cell suppressive activity

Despite the development of severe autoimmune disease, we observed an increase in both the absolute number and percentage of FOXP3+ Treg cells in the lymph nodes (fig. S4A). The expression level of the FOXP3 protein was only slightly altered in Cxxc1-deficient Treg cells (fig. S4B). In an in vitro suppression assay, Treg cells from Foxp3CreCxxc1fl/fl and WT mice exhibited similar suppressive effects on naïve T (Tn) cell proliferation (fig. S4C). The expression of the hallmark Treg-cell marker CTLA-4 showed a modest increase in Cxxc1-deficient Treg cells compared to WT Treg cells, while the expression of GITR remained unchanged (fig. S4D). To further assess the suppressive capacity of Cxxc1-deficient Treg cells in vivo, we employed the experimental autoimmune encephalomyelitis (EAE) model. Naïve CD4+ T cells from 2D2 mice were co-transferred with Treg cells from either Foxp3Cre or Foxp3Cre Cxxc1fl/fl mice into Rag1−/- recipients, and EAE was induced in these recipient mice. Mice that received only naïve CD4+ T cells from 2D2 mice developed more severe EAE symptoms (Fig. 4A). The addition of WT Treg cells from Foxp3Cre mice slightly mitigated EAE progression and reduced Th17 cells in the spinal cord (Fig. 4, A to D). In contrast, Foxp3Cre Cxxc1fl/fl Treg cells failed to suppress EAE (Fig. 4, A to D), and the cKO mice showed a reduction in Treg cells frequency in CNS tissues(Fig. 4E). Finally, we examined the role of CXXC1 in Treg cell-mediated suppression using T cell transfer-induced colitis, in which naïve T cells were transferred to Rag1−/- recipients either alone or together with WT or Foxp3Cre Cxxc1fl/fl Treg cells. The transfer of naïve T cells led to weight loss and intestinal pathology in recipient mice(Fig. 4, F and G). Mice receiving WT Treg cells continued to gain weight (Fig. 4F) , whereas those that received Treg cells from cKO mice were unable to prevent colitis and exhibited a reduced percentage of Treg cells (Fig. 4, F to H). These findings underscore the critical role of CXXC1 in maintaining Treg cell function in vivo.

CXXC1 is essential for Treg cells to suppress T cell-mediated EAE and colitis.

(A)Mean clinical scores for EAE in Rag1−/− recipients of 2D2 CD4+ T cells, either alone or in combination with Foxp3cre or Foxp3creCxxc1fl/fl mice after immunization with MOG35–55, complete Freund’s adjuvant (CFA), and pertussis toxin (n=5). (B and C)Representative histology of the spinal cord of Rag1−/− mice after EAE induction. Hematoxylin and eosin(H&E) staining (upper), Luxol fast blue (F&B) staining (lower). Scale bars, 50 μm (400×) and 200 μm (100×).

(D) Representative flow cytometry plots and quantification of the the percentages of IFNγ+ or IL-17A+ CD4+Vβ11+ T cells (n=5).

(E) Statistical analysis of the percentage CD4+ FOXP3+ Treg cell in CNS tissues 14 days after EAE induction.

(F) Changes in body weight of Rag1−/− mice after colitis induction(n=6).

(G) Haematoxylin and eosin (H&E) staining of colons from T cell-induced colitis mice 6 weeks after T cell transfer. Scale bars, 50 μm (400×) and 200 μm (100×).

(H) Statistical analysis of the percentage CD4+ FOXP3+ Treg cell in the spleen, mesenteric lymph nodes, and colon 6 weeks after colitis induction.

Error bars show mean ± SD. P values are determined by a two-tailed Student’s t-test or two-way ANOVA and Holm–Sidak post hoc test (A, D, E and H). (*P<0.05, **P<0.01, ***PC<C0.001, ****P<0.0001).

Treg cell lineage homeostasis and proliferation depend upon CXXC1

Treg cells harbor a diverse T-cell receptor (TCR) repertoire, which likely play a critical role in their immune suppression function 40–42. To explore the role of CXXC1 in Treg-mediated suppression, we performed single-cell RNA sequencing (scRNA-seq) combined with TCR sequencing (TCR-seq) on CD4+YFP+ Treg cells isolated from mouse lymph nodes. After quality control and removal of doublets, 18,577 cells were retained for further analysis. Through unsupervised clustering and uniform manifold approximation and projection (UMAP) analysis, we identified eight distinct Treg cell clusters based on the expression of well-characterized markers, with a particular focus on two clusters of activated Treg cells that exhibited higher expression of markers and gene sets relative to naïve Treg cells (Fig. 5A and fig. S5, A to C). A comparison between Cxxc1-deficient and WT Treg cells within each cluster revealed a reduction in Cxxc1-deficient cells in the naïve subsets, while an increase was observed in the Gzmb+ and H2-Eb1+ subsets (Fig. 5B). To further elucidate the transition of Treg cells along a dynamic biological timeline, we constructed pseudo-time trajectories using Slingshot 43. The pseudo-time gradient depicted a progression from quiescent to activated Treg cells, ultimately encompassing the Gzmb+ and H2-Eb1+ subsets (Fig. 5C). Given the antigen-specific suppression capabilities of Treg cells 44,45, we examined their clonal expansion. The analysis revealed that expanded WT TCR clonotypes (nC≥C2) were predominantly distributed among the Nt5e+ subsets, while Cxxc1-deficient Treg cells showed expanded clonotypes primarily within the Gzmb+ and H2-Eb1+ subsets(Fig. 5D and Fig. S5D). TCR sharing analysis indicated clonotype sharing among various clusters of WT Treg cells, suggesting a degree of homogeneity. However, the reduced TCR sharing in Cxxc1-deficient Treg cells implies that decreased TCR diversity may impair the suppressive activity of Treg cells (Fig. 5E) 42. Furthermore, the Cxxc1-deficient group exhibited lower expression of several Treg-specific genes associated with suppressive functions, such asNt5e, Il10, Pdcd1, Klrg1, as well as genes that inhibit effector T cells differentiation, including Sell and Tcf7 (Fig. 5F and fig. S5E). Conversely, Cxxc1-deficient Treg cells demonstrated elevated expression of Gzmb, Il2ra, and Cd69 compared to WT, reflecting a profile indicative of increased activation (Fig. 5, F and G). Additionally, we observed increased expression of genes linked to Th1-type inflammation, such as Ifng, Tbx21, and Hif1a, in Cxxc1-deficient Treg cells, likely due to extreme inflammatory conditions (Fig. 5, F to H). The proportion of FOXP3+Ki67+ Treg cells was lower in cKO mice compared to WT mice (Fig. 5I). These findings underscore the crucial role of CXXC1 in in maintaining Treg cell function and homeostasis.

Single-cell transcriptomics reveals distinct Treg cell populations.

(A) UMAP plot showing clusters identified based on variable gene expression of sorted CD4+YFP+ Treg cells. Each dot represents a cell, and each color corresponds to a different population of cell types. Clustering analysis revealed 8 distinct Treg cell populations.

(B) Mean fold changes in cluster abundance between Foxp3cre and Foxp3creCxxc1fl/fl mice. (C)Pseudotime trajectories of Treg cells based on Slingshot, color-coded by Treg cell subpopulations.

(D) Visualization of density and clonotype richness across Treg clusters from Foxp3cre and Foxp3creCxxc1fl/fl mice.

(E) TCR sharing of expanded clonotypes across all possible combinations of Treg cells from Foxp3cre and Foxp3creCxxc1fl/fl mice.

(F) Heatmap showing Z scores for the average expression of Treg-specific genes in each cluster between Foxp3cre and Foxp3creCxxc1fl/fl mice.

(G-H)Representative flow cytometry plots and quantification of (G) expression of CD25, CD69, ICOS, T-bet, and (H) IFN-γ in CD4+YFP+ Treg cells from Foxp3cre and Foxp3creCxxc1fl/fl mice (n=5 CD25, n=10 CD69, n=5 ICOS, n=6 T-bet, n=5 IFN-γ).

(I)Ki-67 expression (left) and frequency (right) in CD4+FOXP3+ Treg cells from Foxp3 cre a-nd Foxp3creCxxc1fl/fl mice (n=6).

Error bars show mean ± SD. P values are determined by a two-tailed Student’s t-test (G-I). (**P<0.01, ***P<0.001, ****P<0.0001). The flow cytometry results are representative of three independent experiments.

Intrinsic Cxxc1 deficiency impairs Treg cell suppression, proliferation, and molecular programs

To confirm that the deficiency in Treg cell function in Treg-specific Cxxc1-deficient animals is due to intrinsic defects caused by Cxxc1 deficiency, rather than severe autoimmune inflammation in Foxp3Cre Cxxc1fl/fl mice, we examined Cxxc1-sufficient and Cxxc1-deficient Treg cell subsets in heterozygous Foxp3Cre/+ Cxxc1fl/fl (designated as “het-KO”) and littermate Foxp3Cre/+ Cxxc1fl/+ (designated as “het-WT”) female mice (Fig. 6A). Notably, het-KO female mice did not exhibit overt signs of autoimmunity, as random X-chromosome inactivation led to the coexistence of both Cxxc1-cKO and Cxxc1-WT Treg cells. However, both the frequency and absolute numbers of FOXP3+YFP+ Treg cells within the total Treg population were reduced in het-KO mice compared to their counterparts in het-WT littermates, indicating that Cxxc1 deficiency imposes a competitive disadvantage on Treg cells (Fig. 6B). Additionally, Cxxc1-deficient YFP+ Treg cells failed to upregulate the proliferation marker Ki-67 (Fig. 6C). Moreover, YFP+ Treg cells in het-KO female mice showed reduced expression of key genes essential for suppressive function, including ICOS, CD25,CTLA4, and GITR, compared to YFP−Treg cells from the same mice (Fig. 6D). Consistently, we confirmed the impaired suppressive function of Treg cells from heterozygous Foxp3Cre/+ Cxxc1fl/fl mice in vitro and in vivo (Fig. 6E and fig. S6A-E). To investigate the molecular program affected by the deletion of Cxxc1 in Treg cells, we performed RNA-sequencing (RNA-seq) analysis on CD4+YFP+ Treg cells isolated from het-WT and het-KO mice. We then conducted a differential gene expression (DGE) analysis based on the RNA-seq data. Among all expressed genes, 865 were upregulated (“Up” genes) and 761 were downregulated (“Down” genes) by ≥1.5-fold in CD4+YFP+ Treg cells from het-KO mice compared to het-WT mice, with a false discovery rate (FDR)–adjusted P-value cutoff of 0.05 (fig. S6F). Gene Ontology (GO) enrichment analysis revealed that the downregulated genes in Cxxc1-deficient Treg cells were predominantly enriched in pathways related to the negative regulation of immune system process and regulation of cell−cell adhesion (fig. S6G). The Cxxc1-deficient Treg cells also showed reduced expression of several genes associated with Treg cell suppressive function, including Il10, Tigit, Lag3, Icos, Nt5e(encoding CD73) and Itgae (encoding CD103) (fig. S6H). Thus, while YFPC WT Treg cells effectively prevent autoimmunity in het-KO mice, the absence of Cxxc1 in YFPC Treg cells disrupts key Treg cell marker expression and impairs their suppressive function under steady-state conditions.

Cxxc1-deficient Treg cells exhibit functional impairment and disrupted homeostasis in steady-state conditions.

(A) Schematic representation of wild-type and Cre-positive Treg cells in female Foxp3Cre/+mice.

(B) Flow cytometry analysis of the YFP−FOXP3+ (WT) and YFP+FOXP3+ (KO) Treg cells in Foxp3Cre/+ Cxxc1fl/+ (het-WT) and Foxp3Cre/+ Cxxc1fl/fl ( het-KO) female mice (left), along with the frequency and absolute numbers of YFP+ cells within the total Treg population (right)(n=5).

(C) Flow cytometry analysis of Ki-67expression (left) and MFI (right) in YFP− and YFP+ cells within the CD4+FOXP3+ Treg cells from 6- to 8-week-old het-KO female mice (n=4).

(D)Representative flow cytometry plots and quantification of ICOS, CD25, CTLA4, and GITR expression in YFP− and YFP+ cells within CD4+FOXP3+ Treg cells from het-KO female mice (n=5).

(E)Suppression of CFSE-labelled Tn cell proliferation by different ratios of CD4+YFP+ Treg cells from Foxp3Cre/+ Cxxc1fl/+ and Foxp3Cre/+ Cxxc1fl/fl female mice. On the right, the percentage of proliferated responding T cells is presented (n=5).

Error bars show mean ± SD. P values are determined by a two-tailed Student’s t-test (B-E). (ns, not significant. *P<0.05,**P<0.01, ***PC<C0.001, ****P<0.0001). The flow cytometry results are representative of three independent experiments.

The FOXP3-CXXC1 complex regulates the expression of key factors in Treg cells that are associated with the breadth of H3K4me3

CXXC1 binds to unmethylated CpG DNA via its N-terminal CXXC finger domain, facilitating its interaction with DNA methyltransferase 1 (DNMT1). This binding stabilizes the DNMT1 protein, thereby regulating DNA methylation 46,47. To investigate whether CXXC1 depletion affects DNA methylation in Treg cells, we performed whole genome bisulfite sequencing (WGBS) on Treg cells isolated from both WT and cKO mice. On average, Cxxc1-deficient Treg cells exhibited no changes in DNA methylation at gene loci or across genome-wide CpG sites, irrespective of chromosomal region (fig. S7, A to C). Furthermore, Cxxc1 knockout Treg cells did not show a increase in DNA methylation at key Treg signature gene loci (fig. S7D). Given the pivotal role of MLL4-mediated H3K4me1 in establishing the enhancer landscape and facilitating long-range chromatin interactions during Treg cell development15, we performed CUT&Tag to assess changes in H3K4me1 levels in Cxxc1-deficient Treg cells. This analysis revealed that H3K4me1 levels were similar in both WT and Cxxc1-deficient Treg cells (fig. S7, E to G).

While H3K4me3 modifications typically form sharp 1-to 2-kb peaks around promoters, some genes exhibit broader H3K4me3 regions, referred to as broad H3K4me3 domains (H3K4me3-BDs), which can extend to cover part or all of the gene’s coding sequences (up to 20 kb) 48,49. Broad H3K4me3 domains are preferentially associated with genes essential for the identity or function of specific cell types 48,50 and have been implicated in enhancing transcriptional elongation and increasing enhancer activity 50. To further explore the connection between broad H3K4me3 domains and the expression of immune-regulatory genes, we analyzed genes containing broad H3K4me3 regions. We classified the H3K4me3 domains surrounding transcription start sites (TSSs) into three categories: broad (more than 5 kb), medium (between 1 and 5 kb), and narrow (less than 1 kb) (Fig. 7A). Notably, Cxxc1-deficient Treg cells exhibited weaker H3K4me3 signals compared to WT cells within the broad H3K4me3 domains where CXXC1 binding is prominent (Fig. 7, A and B). Using the criteria established by Benayoun et al.48, which defines the top 5% of the widest H3K4me3 domains as BDs, we observed similar enrichment results (fig. S7, H and I). We then compared three groups of genes: BD-associated genes with reduced H3K4me3 levels following Cxxc1 deletion, genes with direct CXXC1 binding, and genes with direct FOXP3 binding. The Venn diagram revealed that the majority of genes (283 out of 294, 96%) with CXXC1 binding and reduced H3K4me3 levels overlap with FOXP3-bound genes, suggesting that CXXC1 presence enhances FOXP3 binding within the broad H3K4me3 domain (Fig. 7C). Furthermore, GO term analysis indicated that BD-associated genes are enriched in biological processes related to the negative regulation of immune system processes (Fig. 7D). Genome browser views displayed the enrichments of FOXP3, CXXC1, and H3K4me3 at key signature genes in Treg cells, such as Ctla4, Il2ra, Icos, and Tnfrsf18, with lower H3K4me3 densities observed at these loci in Cxxc1-deficient Treg cells (Fig. 7E). Similar patterns were observed at core genes involved in Treg homeostasis and suppressive function (e.g., Lag3, Nt5e, Ikzf4, and Cd28) (fig. S7, J)51,52. These findings suggest that CXXC1 and FOXP3 collaboratively promote sustained Treg cell homeostasis and function by preserving the H3K4me3 modification at key Treg cell genes.

CXXC1 regulates Foxp3-dependent molecule H3K4 trimethylation in Treg cells.

(A-B)Heatmaps showing H3K4me3(A)and CXXC1(B) signals centered on narrow, medium, and broad domains. The top panel shows the average CUT&Tag signals around indicated domains.

(D) Venn diagrams showing the overlap of FOXP3 binding genes, CXXC1 binding genes, and H3K4me3-BD-associated genes with decreased H3K4me3 levels after CXXC1 depletion in Treg cells.

(E) Gene Ontology (GO) pathway analysis of the overlap peaks.

(F) Representative genome browser view showing the enrichments of FOXP3, CXXC1, and H3K4me3 in Treg cells.

We previously demonstrated that the FOXP3-CXXC1 complex plays a key role in modulating H3K4me3 deposition at Treg-specific gene loci. To further clarify whether Cxxc1 deletion affects FOXP3 binding to its target genes, we performed CUT&Tag experiments to compare FOXP3 binding profiles between WT and Cxxc1 KO Treg cells. The results revealed that most FOXP3-bound regions in WT Treg cells were similarly enriched in KO Treg cells, indicating that Cxxc1 deletion does not impair FOXP3’s DNA-binding ability (fig. S8A and B). Together, these findings suggest that the regulatory role of CXXC1 in Treg cells is mediated through its effect on H3K4me3 deposition rather than altering FOXP3’s binding to DNA.

Discussion

Treg cells specifically express the transcription factor FOXP3, which is essential for maintaining Treg lineage stability and suppressive function53. However, FOXP3 alone is insufficient to fully regulate the transcriptional signature and functionality of Treg cells; its interaction with protein partners is crucial for this regulation. In this study, we identify CXXC1 as a critical epigenetic regulator and functional cofactor of FOXP3. Although CXXC1 is not required for FOXP3’s DNA-binding activity, as evidenced by similar FOXP3 binding patterns in WT and Cxxc1-deficient Treg cells (CUT&Tag analysis), it plays an essential role in maintaining H3K4me3 modifications at FOXP3 target loci. By acting as both an epigenetic regulator and a FOXP3 cofactor, CXXC1 ensures the stability of the Treg transcriptional program, highlighting its pivotal role in preserving Treg cell functionality and immune homeostasis.

FOXP3 is known to promote histone H3 acetylation at the promoters and enhancers of its target genes, such as Il2ra, Ctla4, and Tnfrsf18, following Treg cell activation, thereby functioning as a transcriptional activator54,55. Conversely, FOXP3 can act as a repressor by silencing target genes like Il2 and Ifng through the induction of histone H3 deacetylation, mediated by the recruitment of histone deacetylases and transcriptional co-repressors10,56. Additionally, FOXP3 exerts this repression by recruiting the Ezh2-containing polycomb repressive complex to target genes during activation, as FOXP3-repressed genes are associated with H3K27me3 deposition and reduced chromatin accessibility11,57. Recent studies have begun to explore the biological significance of H3K4me3 breadth, revealing a positive correlation between H3K4me3 breadth and gene expression 50,58,59. These studies also suggest that H3K4me3 breadth contributes to defining specific cell identities during development and disease, including systemic autoimmune diseases like systemic lupus erythematosus and various cancers 48,60–63. Our study demonstrates that the loss of Cxxc1 leads to reduced H3K4me3 levels, predominantly in genes with broader peaks, such as Il2ra, Tnfrsf18, Ctla4 and Icos, directly impairing their suppressive function in Treg cells. Although our research primarily focused on the role of CXXC1 in Treg cells, it is plausible that similar mechanisms may be operative in other cell types.

The Treg-specific deletion of Cxxc1 leads to a rapid and fatal autoimmune disorder, characterized by systemic inflammation and tissue damage, underscoring the essential role of CXXC1 in maintaining immune self-tolerance within Treg cells. Interestingly, despite this severe phenotype, our findings show that H3K4me1 levels were comparable between WT and Cxxc1-deficient Treg cells. In contrast, MLL4 is critical for Treg cell development in the thymus, primarily by regulating H3K4me1, though it is not required for peripheral Treg cell function15. Numerous studies investigating the effectors of Treg cell-mediated suppression have identified a wide range of molecules and mechanisms. These include the upregulation of the inhibitory co-stimulatory receptor CTLA-4, which initiates inhibitory signaling; the sequestration of the T-cell growth factor IL-2 via CD25; the secretion of inhibitory cytokines, such as interleukin (IL)-10, IL-35 and TGF-β and the activity of ectoenzymes CD39 and CD73 on the Treg cells surface, which convert extracellular ATP, a potent pro-inflammatory mediator, into its anti-inflammatory counterpart, adenosine42,64. Mechanistically, we demonstrated that CXXC1 interacts with FOXP3 to regulate Treg cell function by trimethylating H3K4 at broad H3K4me3 domains of multiple genes involved in suppressive functions such as Il2ra, Nt5e, and Ctla4. Additionally, MLL1, another KMT, controls Treg cell activation and function by specifically regulating H3K4 trimethylation at genes encoding key Treg-related molecules such as Tigit, Klrg1, Tbx21, Cxcr3, and serves as a crucial epigenetic regulator in establishing a stable Th1-Treg lineage14.

Our findings demonstrate that Cxxc1-deficient Treg cells exhibit reduced H3K4me3 levels at Treg-specific loci, indicating that CXXC1 plays a critical role in regulating this epigenetic modification. This is consistent with previous studies showing that CXXC1 acts as a non-catalytic component of the Set1/COMPASS complex 65–68, which includes the H3K4 methyltransferases SETD1A and SETD1B, the primary enzymes responsible for H3K4 trimethylation. Given this, we propose that CXXC1 supports H3K4me3 deposition in Treg cells by interacting with and stabilizing the activity of the Set1/COMPASS complex. Further studies are required to directly investigate the interactions between CXXC1 and these methyltransferases in the Treg cell context.

Our results offer provide novel insights into the suppressive functions, heterogeneity, and regulatory mechanisms of Treg cells. Maintaining Treg cell homeostasis and function remains a challenge in harnessing Treg cells for the treatment of autoimmune diseases and the prevention of graft rejection.